---
title: "COTS Fine-Scale SDM: Comprehensive Validation Report"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    code-fold: true
    theme: cosmo
    fig-width: 10
    fig-height: 8
  pdf:
    toc: true
execute:
  warning: false
  message: false
---

## Overview

This report validates the XGBoost COTS Fine-Scale Species Distribution Model using historical Manta Tow and Cull data. We compare both the **2025 Year-Specific** and **Year-Agnostic** models.

The validation involves:
1. **Grid-Based Historical Validation (100m)**: General statistics (ROC, AUC, Confusion Matrix) evaluating the model against historical COTS/scar presence.
2. **Tow-Path Simulation**: Operational simulation showing which manta tows would be suggested based on the model.
3. **Cull Dive Financial Savings**: Quantification of dive hours and dollars saved by avoiding ineffective culls (0.00 or <0.02 CPUE).
4. **Reef Maps**: Visual mapping of predictions vs. operations.

```{r setup}
library(terra)
library(sf)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
library(tidyr)
library(knitr)
library(stringr)
library(pROC)
library(tune)
library(yardstick)

# --- Configuration ---
predict_agnostic_path    <- "outputs/COTS_prob_0.02_cpue_agnostic.tif"
predict_year2025_path    <- "outputs/COTS_prob_0.02_cpue_year2025_clean.tif"  # native 30m raster
predict_year2025_rg_path <- "outputs/COTS_prob_0.02_cpue_year2025_withRG.tif"
predict_manta_clean_path <- "outputs/COTS_prob_manta_clean.tif"
predict_manta_rg_path    <- "outputs/COTS_prob_manta_withRG.tif"
predict_maxent_path      <- "outputs/COTS_maxent_suitability_clean.tif"
predict_maxent_rg_path   <- "outputs/COTS_maxent_suitability_reefguide.tif"
predict_ensemble_path    <- "outputs/COTS_prob_ensemble.tif"

manta_csv <- "data/COTS Program  Manta Tow Data-2026-02-04.csv"
cull_xlsx <- "data/250929_COTS-Manta-Cull-RHIS-Data-Matthews-and-Schlawinsky.xlsx"
cull_polys_gpkg <- "data/cull_site_polygons.gpkg"

out_dir <- "validation_outputs"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Selected Reefs
reef_select <- c(
  "Tongue Reef (16-026)",
  "Batt Reef (16-029)",
  "Arlington Reef (16-064)",
  "(Big) Broadhurst Reef (No 1) (18-100a)",
  "Britomart Reef (18-024)",
  "Sudbury Reef (No 1) (17-001a)",
  "Trunk Reef (18-027)",
  "Bramble Reef (18-029)",
  "Davies Reef (18-096)",
  "Heron Reef (23-052a)"
)
```

## 1. Load and Prepare Data

```{r load_data}
# Load Model Predictions
prob_agnostic    <- if(file.exists(predict_agnostic_path))    terra::rast(predict_agnostic_path) else NULL
prob_year2025    <- if(file.exists(predict_year2025_path))    terra::rast(predict_year2025_path) else NULL
prob_2025_rg     <- if(file.exists(predict_year2025_rg_path)) terra::rast(predict_year2025_rg_path) else NULL
prob_manta_clean <- if(file.exists(predict_manta_clean_path)) terra::rast(predict_manta_clean_path) else NULL
prob_manta_rg    <- if(file.exists(predict_manta_rg_path))    terra::rast(predict_manta_rg_path) else NULL
prob_maxent      <- if(file.exists(predict_maxent_path))      terra::rast(predict_maxent_path) else NULL
prob_maxent_rg   <- if(file.exists(predict_maxent_rg_path))   terra::rast(predict_maxent_rg_path) else NULL
prob_ensemble    <- if(file.exists(predict_ensemble_path))    terra::rast(predict_ensemble_path) else NULL

if(is.null(prob_year2025) || is.null(prob_agnostic)) stop("Missing core prediction rasters.")

# Load Manta Tow Data
manta_raw <- read_csv(manta_csv, show_col_types = FALSE) %>%
  filter(
    !is.na(StartLongitude), !is.na(StartLatitude), !is.na(EndLongitude), !is.na(EndLatitude),
    StartLongitude != 0, StartLatitude != 0, EndLongitude != 0, EndLatitude != 0
  ) %>%
  mutate(
    cots_observed = CrownOfThornsStarfishCount > 0,
    scars_observed = !is.na(ScarsCount) & ScarsCount > 0,
    presence_any = cots_observed | scars_observed,
    ReefName = as.character(ReefName)
  )

# Convert Manta Tows to spatial lines
make_tow_lines <- function(df) {
  lines <- lapply(seq_len(nrow(df)), function(i) {
    p1 <- c(df$StartLongitude[i], df$StartLatitude[i])
    p2 <- c(df$EndLongitude[i], df$EndLatitude[i])
    st_linestring(rbind(p1, p2))
  })
  st_sf(df, geometry = st_sfc(lines, crs = 4326))
}

manta_sf <- make_tow_lines(manta_raw)
manta_sf <- st_transform(manta_sf, crs = crs(prob_year2025))
manta_selected <- manta_sf %>% filter(ReefName %in% reef_select)

# Load Cull Data
cull_raw <- read_excel(cull_xlsx, sheet = 4) %>%
  filter(!is.na(Longitude), !is.na(Latitude), !is.na(Bottomtime), Bottomtime > 0, Longitude != 0, Latitude != 0) %>%
  filter(as.Date(SurveyDate) > as.Date("2018-11-30")) %>%
  mutate(
    Total = Cohort1 + Cohort2 + Cohort3 + Cohort4,
    CPUE = Total / Bottomtime,
    is_ghost = (Total == 0),
    is_non_problematic = (CPUE < 0.02),
    ReefName = as.character(ReefName)
  )

# Top 30 reefs with most culling hours in 2025
reef_select_eval <- cull_raw %>% 
  filter(as.Date(SurveyDate) >= as.Date("2025-01-01")) %>%
  group_by(ReefName) %>%
  summarise(Total_Hours = sum(Bottomtime, na.rm=TRUE) / 60) %>%
  arrange(desc(Total_Hours)) %>%
  slice_head(n = 30) %>%
  pull(ReefName)

# Manta tows for the selected 10 mapping reefs AND the 30 evaluated reefs
manta_selected <- manta_sf %>% filter(ReefName %in% reef_select)
manta_eval_all <- manta_sf %>% filter(ReefName %in% reef_select_eval)

# Cull dives for the selected 10 mapping reefs
cull_selected <- cull_raw %>% filter(ReefName %in% reef_select)
cull_sf <- st_as_sf(cull_selected, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = crs(prob_year2025))

# Cull dives for the 30 evaluated reefs (point geometry)
cull_sf_eval_all <- st_as_sf(cull_raw %>% filter(ReefName %in% reef_select_eval), coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = crs(prob_year2025))

# Load cull site polygons and join to cull data
cull_site_polys <- st_read(cull_polys_gpkg, quiet = TRUE) %>%
  st_transform(crs = crs(prob_year2025))

cat("Loaded", nrow(cull_site_polys), "cull site polygons from gpkg\n")
cat("Matching to", n_distinct(cull_raw$CullSiteName), "unique cull site names...\n")
cat("Match rate:", round(mean(cull_raw$CullSiteName %in% cull_site_polys$CullSiteName) * 100, 1), "%\n")

# Calculate Default Threshold: Median of 2025 raster cropped to selected reefs
ext_val <- terra::ext(manta_selected) + 2000
crop_2025 <- terra::crop(prob_year2025, ext_val)
median_thresh_2025 <- as.numeric(terra::global(crop_2025, fun=function(x) median(x, na.rm=TRUE)))

crop_agnostic <- terra::crop(prob_agnostic, ext_val)
median_thresh_agnostic <- as.numeric(terra::global(crop_agnostic, fun=function(x) median(x, na.rm=TRUE)))
crop_maxent <- if(!is.null(prob_maxent)) terra::crop(prob_maxent, ext_val) else NULL
median_thresh_maxent <- if(!is.null(crop_maxent)) as.numeric(terra::global(crop_maxent, fun=function(x) median(x, na.rm=TRUE))) else NA
thresh_maxent_p88    <- if(!is.null(crop_maxent)) as.numeric(terra::global(crop_maxent, fun=function(x) quantile(x, probs=0.88, na.rm=TRUE))) else NA
thresh_maxent_p95    <- if(!is.null(crop_maxent)) as.numeric(terra::global(crop_maxent, fun=function(x) quantile(x, probs=0.95, na.rm=TRUE))) else NA

# Calculate P88 and P95 Thresholds
thresh_2025_p88 <- as.numeric(terra::global(crop_2025, fun=function(x) quantile(x, probs=0.88, na.rm=TRUE)))
thresh_2025_p95 <- as.numeric(terra::global(crop_2025, fun=function(x) quantile(x, probs=0.95, na.rm=TRUE)))

# Fixed mapping cutoffs
p_min_2025 <- 0.48
p_max_2025 <- 0.55

cat("Median Threshold (2025):", round(median_thresh_2025, 3), "\n")
cat("Median Threshold (Agnostic):", round(median_thresh_agnostic, 3), "\n")
cat("Median Threshold (MaxEnt):", round(median_thresh_maxent, 3), "\n")
cat("88th Percentile (MaxEnt):", round(thresh_maxent_p88, 3), "\n")
cat("95th Percentile (MaxEnt):", round(thresh_maxent_p95, 3), "\n")
cat("88th Percentile (2025):", round(thresh_2025_p88, 3), "\n")
cat("95th Percentile (2025):", round(thresh_2025_p95, 3), "\n")
cat("Mapping Range (2025):", p_min_2025, "to", p_max_2025, "\n")
```


## 1. Model Fitting Performance

This section presents the internal performance metrics generated during the model training process. The Cull Dive XGBoost models were trained using 5-fold Spatial Block Cross-Validation (50km blocks) on historical cull dive data, while the Ensemble Meta-Learners were evaluated via Spatial Out-of-Fold (OOF) cross-validation across reefs.

- **Training (In-Fold)** represents the model's performance when tested on the data it was trained on.
- **Validation (Out-of-Fold)** represents the model's performance on hold-out spatial blocks, providing a robust estimate of how models generalize to un-surveyed reefs.

```{r model_fitting_metrics, results='asis'}
library(pROC)
library(yardstick)

metrics_clean_file <- "outputs/fitted_model_clean.rds"
metrics_rg_file    <- "outputs/fitted_model_withRG.rds"
metrics_ens_file   <- "outputs/fitted_ensemble_model.rds"

compute_detailed_metrics <- function(truth_vec, prob_vec, pred_class_vec = NULL, model_name, dataset_name, threshold = 0.5) {
  truth_b <- as.character(truth_vec) %in% c("1", "TRUE", "true")
  if (is.null(pred_class_vec)) {
    pred_b <- prob_vec >= threshold
  } else {
    pred_b <- as.character(pred_class_vec) %in% c("1", "TRUE", "true")
  }
  
  roc_obj <- pROC::roc(response = truth_b, predictor = prob_vec, quiet = TRUE)
  auc_val <- as.numeric(pROC::auc(roc_obj))
  
  # Confusion Matrix components
  # Class 1 = Successful dive (CPUE >= 0.02)
  # Class 0 = Failed dive (CPUE < 0.02)
  tp <- sum(pred_b & truth_b)
  fp <- sum(pred_b & !truth_b)
  fn <- sum(!pred_b & truth_b)
  tn <- sum(!pred_b & !truth_b)
  
  total <- tp + fp + fn + tn
  accuracy <- (tp + tn) / total
  
  # Metrics for Successful Dives (CPUE >= 0.02)
  prec_succ <- ifelse(tp + fp > 0, tp / (tp + fp), NA)
  rec_succ  <- ifelse(tp + fn > 0, tp / (tp + fn), NA) # Sensitivity
  f1_succ   <- ifelse(!is.na(prec_succ) && !is.na(rec_succ) && (prec_succ + rec_succ > 0), 
                      2 * (prec_succ * rec_succ) / (prec_succ + rec_succ), NA)
  
  # Metrics for Failed Dives (CPUE < 0.02)
  prec_fail <- ifelse(tn + fn > 0, tn / (tn + fn), NA)
  rec_fail  <- ifelse(tn + fp > 0, tn / (tn + fp), NA) # Specificity
  f1_fail   <- ifelse(!is.na(prec_fail) && !is.na(rec_fail) && (prec_fail + rec_fail > 0), 
                      2 * (prec_fail * rec_fail) / (prec_fail + rec_fail), NA)
  
  tibble(
    Model = model_name,
    Dataset = dataset_name,
    AUC = round(auc_val, 3),
    Accuracy = round(accuracy, 3),
    `Prec (>0.02)` = round(prec_succ, 3),
    `Rec (>0.02)`  = round(rec_succ, 3),
    `F1 (>0.02)`   = round(f1_succ, 3),
    `Prec (<0.02)` = round(prec_fail, 3),
    `Rec (<0.02)`  = round(rec_fail, 3),
    `F1 (<0.02)`   = round(f1_fail, 3)
  )
}

if (file.exists(metrics_clean_file) && file.exists(metrics_rg_file)) {
  m_clean <- readRDS(metrics_clean_file)
  m_rg    <- readRDS(metrics_rg_file)
  
  # 1. Spatial Block Cross-Validation (Out-of-Fold)
  b_clean <- select_best(m_clean$cv_res, metric = "roc_auc")
  cv_clean <- collect_predictions(m_clean$cv_res) %>% filter(.config == b_clean$.config)
  b_rg <- select_best(m_rg$cv_res, metric = "roc_auc")
  cv_rg <- collect_predictions(m_rg$cv_res) %>% filter(.config == b_rg$.config)

  m_cv_clean <- compute_detailed_metrics(cv_clean$cots_problem, cv_clean$.pred_1, cv_clean$.pred_class, "Cull Dive Clean (12 Predictors)", "Cross-Validation (Out-of-Fold)")
  m_cv_rg    <- compute_detailed_metrics(cv_rg$cots_problem, cv_rg$.pred_1, cv_rg$.pred_class, "Cull Dive Reef Guide (17 Predictors)", "Cross-Validation (Out-of-Fold)")

  cv_list <- list(m_cv_clean, m_cv_rg)

  if (file.exists(metrics_ens_file)) {
    m_ens <- readRDS(metrics_ens_file)
    cv_list[[length(cv_list) + 1]] <- tibble(
      Model = "Performance-Weighted Ensemble",
      Dataset = "Cross-Validation (Out-of-Fold)",
      AUC = round(m_ens$auc_weighted, 3),
      Accuracy = NA, `Prec (>0.02)` = NA, `Rec (>0.02)` = NA, `F1 (>0.02)` = NA,
      `Prec (<0.02)` = NA, `Rec (<0.02)` = NA, `F1 (<0.02)` = NA
    )
    cv_list[[length(cv_list) + 1]] <- tibble(
      Model = "Stacked Logistic Meta-Learner",
      Dataset = "Cross-Validation (Spatial OOF)",
      AUC = round(m_ens$auc_logistic_oof, 3),
      Accuracy = NA, `Prec (>0.02)` = NA, `Rec (>0.02)` = NA, `F1 (>0.02)` = NA,
      `Prec (<0.02)` = NA, `Rec (<0.02)` = NA, `F1 (<0.02)` = NA
    )
    cv_list[[length(cv_list) + 1]] <- tibble(
      Model = "Stacked XGBoost Meta-Learner",
      Dataset = "Cross-Validation (Spatial OOF)",
      AUC = round(m_ens$auc_xgb_oof, 3),
      Accuracy = NA, `Prec (>0.02)` = NA, `Rec (>0.02)` = NA, `F1 (>0.02)` = NA,
      `Prec (<0.02)` = NA, `Rec (<0.02)` = NA, `F1 (<0.02)` = NA
    )
  }

  cat("### 1.1 Comparative Model Performance: Spatial Cross-Validation\n\n")
  cv_table <- bind_rows(cv_list)
  print(kable(cv_table, caption = "Cross-Validation Evaluation Metrics: Precision, Recall & F1 for Successful (>=0.02) and Failed (<0.02) Dives"))
  cat("\n\n")
  
  # 2. Regional Metrics Table (Clean Model)
  if (file.exists("outputs/model_performance_metrics_clean.rds")) {
    m_clean_metrics <- readRDS("outputs/model_performance_metrics_clean.rds")
    cat("### 1.2 Regional Validation Performance (Cull Dive Clean Model Out-of-Fold)\n\n")
    print(kable(m_clean_metrics$By_Zone %>% arrange(ManagementZone), digits = 3, caption = "Cull Dive Clean Model Performance by Management Zone"))
    cat("\n\n")
  }
} else {
  cat("*Model performance files not found. Ensure `fit_predict_cots_sdm.R` has been run recently.*")
}
```

### 1.3 MaxEnt Presence-Only Models (Clean vs Reef Guide ENMeval Tuning — 10,000 Background Points)

```{r maxent_eval_summary, results='asis'}
maxent_clean_rds <- ifelse(file.exists("outputs/maxent_enmeval_results_10k.rds"), "outputs/maxent_enmeval_results_10k.rds", "outputs/maxent_enmeval_results.rds")
maxent_rg_rds    <- ifelse(file.exists("outputs/maxent_enmeval_results_reefguide_10k.rds"), "outputs/maxent_enmeval_results_reefguide_10k.rds", "outputs/maxent_enmeval_results_reefguide.rds")

if (file.exists(maxent_clean_rds)) {
  me_res_clean <- readRDS(maxent_clean_rds)
  res_clean_df <- me_res_clean@results %>%
    select(fc, rm, auc.train, auc.val.avg, auc.val.sd, AICc) %>%
    arrange(AICc) %>%
    mutate(Model = "MaxEnt Clean (12-pred, 10k bg)")
    
  best_clean <- res_clean_df[1, ]
  
  cat(paste0("**Optimal MaxEnt Clean Configuration (10,000 Background Points)**:\n",
             "- **Feature Classes**: ", best_clean$fc, " | **rm**: ", best_clean$rm, " | **AICc**: ", round(best_clean$AICc, 2), " | **Validation AUC**: ", round(best_clean$auc.val.avg, 3), " ± ", round(best_clean$auc.val.sd, 3), "\n\n"))
  
  if (file.exists(maxent_rg_rds)) {
    me_res_rg <- readRDS(maxent_rg_rds)
    res_rg_df <- me_res_rg@results %>%
      select(fc, rm, auc.train, auc.val.avg, auc.val.sd, AICc) %>%
      arrange(AICc) %>%
      mutate(Model = "MaxEnt Reef Guide (17-pred, 10k bg)")
      
    best_rg <- res_rg_df[1, ]
    cat(paste0("**Optimal MaxEnt Reef Guide Configuration (10,000 Background Points)**:\n",
               "- **Feature Classes**: ", best_rg$fc, " | **rm**: ", best_rg$rm, " | **AICc**: ", round(best_rg$AICc, 2), " | **Validation AUC**: ", round(best_rg$auc.val.avg, 3), " ± ", round(best_rg$auc.val.sd, 3), "\n\n"))
    
    maxent_comp <- bind_rows(head(res_clean_df, 5), head(res_rg_df, 5))
    print(kable(maxent_comp, digits = 3, caption = "Top Candidate MaxEnt Configurations (Clean vs Reef Guide — 10k Background Points)"))
    cat("\n\n")
  }
} else {
  cat("*MaxEnt ENMeval tuning results not found.*")
}
```

```{r maxent_plots, echo=FALSE, fig.show='hold', out.width='48%'}
vip_clean_img <- ifelse(file.exists("outputs/maxent_vip_cots_10k.png"), "outputs/maxent_vip_cots_10k.png", "outputs/maxent_vip_cots.png")
vip_rg_img    <- ifelse(file.exists("outputs/maxent_vip_reefguide_10k.png"), "outputs/maxent_vip_reefguide_10k.png", "outputs/maxent_vip_reefguide.png")

if(file.exists(vip_clean_img)) knitr::include_graphics(vip_clean_img)
if(file.exists(vip_rg_img))    knitr::include_graphics(vip_rg_img)
```

## 2. Step A: Grid-Based Historical Validation (100m)

We rasterize historical manta tow observations onto a 100m grid and compare them to the SDM probability values (using both 2025 and Agnostic models). We calculate ROC curves, AUC, and evaluate thresholds using confusion matrices.

```{r step_a_grid, fig.width=12, fig.height=6}
if(nrow(manta_selected) > 0) {
  # crop_2025 is already 60m. Resample crop_agnostic to match it exactly.
  pred_60m_2025 <- crop_2025
  pred_60m_agn <- terra::resample(crop_agnostic, crop_2025, method = "bilinear")
  pred_60m_max <- if(!is.null(crop_maxent)) terra::resample(crop_maxent, crop_2025, method = "bilinear") else NULL
  
  # Rasterize Manta Tow presence (1 if ever present, 0 if absent)
  manta_presence <- terra::rasterize(
    terra::vect(manta_selected), 
    pred_60m_2025, 
    field = "presence_any", 
    fun = "max", 
    background = NA
  )
  
  # Build evaluation dataframe
  val_df <- data.frame(
    observed_presence = terra::values(manta_presence)[,1],
    prob_2025 = terra::values(pred_60m_2025)[,1],
    prob_agnostic = terra::values(pred_60m_agn)[,1],
    prob_maxent   = if(!is.null(pred_60m_max)) terra::values(pred_60m_max)[,1] else NA
  ) %>% drop_na()
  
  # --- ROC and AUC ---
  roc_2025 <- roc(val_df$observed_presence, val_df$prob_2025, quiet=TRUE)
  roc_agn <- roc(val_df$observed_presence, val_df$prob_agnostic, quiet=TRUE)
  roc_maxent <- if("prob_maxent" %in% names(val_df)) roc(val_df$observed_presence, val_df$prob_maxent, quiet=TRUE) else NULL
  
  cat("AUC (2025 Model):", round(auc(roc_2025), 3), "\n")
  cat("AUC (Agnostic Model):", round(auc(roc_agn), 3), "\n\n")
  if(!is.null(roc_maxent)) cat("AUC (MaxEnt Model):", round(auc(roc_maxent), 3), "\n")
  
  # Plot ROC
  par(mfrow=c(1,2))
  plot(roc_2025, main="ROC Curve (2025)", col="blue", lwd=2)
  plot(roc_agn, main="ROC Curve (Agnostic)", col="red", lwd=2)
  
  # --- Confusion Matrix at Median Threshold ---
  cat("Confusion Matrix for 2025 Model (Threshold = Median:", round(median_thresh_2025,3), ")\n")
  cm_2025 <- table(
    Predicted = factor(val_df$prob_2025 >= median_thresh_2025, levels=c(FALSE, TRUE)),
    Observed = factor(val_df$observed_presence == 1, levels=c(FALSE, TRUE))
  )
  print(cm_2025)
  cat("Accuracy:", sum(diag(cm_2025))/sum(cm_2025), "\n\n")
  
  cat("Confusion Matrix for Agnostic Model (Threshold = Median:", round(median_thresh_agnostic,3), ")\n")
  cm_agn <- table(
    Predicted = factor(val_df$prob_agnostic >= median_thresh_agnostic, levels=c(FALSE, TRUE)),
    Observed = factor(val_df$observed_presence == 1, levels=c(FALSE, TRUE))
  )
  print(cm_agn)
  cat("Accuracy:", sum(diag(cm_agn))/sum(cm_agn), "\n")
}
```

## 2b. Cull Dive Validation (ROC & AUC across Base & Ensemble SDMs)

Cull dive data provides a highly reliable validation target because divers directly count COTS at each site. We extract spatial predictions at each cull dive location (Jan 2025 onwards across top 30 evaluated reefs) and compare model discrimination and classification performance across all base model configurations and the Performance-Weighted Ensemble:
1. **Ensemble SDM (`COTS_prob_ensemble.tif`)**
2. **Cull Dive XGBoost Reef Guide (17-pred)**
3. **Cull Dive XGBoost Clean (12-pred)**
4. **MaxEnt Reef Guide (17-pred, 10k bg)**
5. **MaxEnt Clean (12-pred, 10k bg)**
6. **Manta Tow XGBoost Clean (12-pred)**
7. **Manta Tow XGBoost Reef Guide (17-pred)**

```{r step_a_cull_roc, fig.width=12, fig.height=10}
# Prepare Cull Dive point locations
cull_pts_sub <- cull_raw %>% 
  filter(ReefName %in% reef_select_eval, as.Date(SurveyDate) >= as.Date("2025-01-01"))

cull_pts_sf <- st_as_sf(cull_pts_sub, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = crs(prob_year2025))

v_cull_pts <- terra::vect(cull_pts_sf)

# Extract predictions for all model variants
cull_val_df <- tibble(
  observed_presence = as.integer(pull(cull_pts_sub, CPUE) >= 0.02),
  CPUE              = pull(cull_pts_sub, CPUE),
  cull_clean  = if(!is.null(prob_year2025))    as.numeric(terra::extract(prob_year2025, v_cull_pts)[[2]]) else NA_real_,
  cull_rg     = if(!is.null(prob_2025_rg))     as.numeric(terra::extract(prob_2025_rg, v_cull_pts)[[2]]) else NA_real_,
  manta_clean = if(!is.null(prob_manta_clean)) as.numeric(terra::extract(prob_manta_clean, v_cull_pts)[[2]]) else NA_real_,
  manta_rg    = if(!is.null(prob_manta_rg))    as.numeric(terra::extract(prob_manta_rg, v_cull_pts)[[2]]) else NA_real_,
  maxent_clean= if(!is.null(prob_maxent))      as.numeric(terra::extract(prob_maxent, v_cull_pts)[[2]]) else NA_real_,
  maxent_rg   = if(!is.null(prob_maxent_rg))   as.numeric(terra::extract(prob_maxent_rg, v_cull_pts)[[2]]) else NA_real_,
  ensemble    = if(!is.null(prob_ensemble))    as.numeric(terra::extract(prob_ensemble, v_cull_pts)[[2]]) else NA_real_
) %>% drop_na()

cat("Cull dive validation dataset:", nrow(cull_val_df), "dives\n\n")

# --- Calculate ROC curves ---
y_true <- pull(cull_val_df, observed_presence)

roc_models <- list(
  "Ensemble SDM (Weighted)"              = pull(cull_val_df, ensemble),
  "Cull Dive XGBoost Reef Guide (17-pred)"= pull(cull_val_df, cull_rg),
  "Cull Dive XGBoost Clean (12-pred)"     = pull(cull_val_df, cull_clean),
  "MaxEnt Reef Guide (17-pred, 10k)"    = pull(cull_val_df, maxent_rg),
  "MaxEnt Clean (12-pred, 10k)"         = pull(cull_val_df, maxent_clean),
  "Manta XGBoost Clean (12-pred)"       = pull(cull_val_df, manta_clean),
  "Manta XGBoost Reef Guide (17-pred)"  = pull(cull_val_df, manta_rg)
)

roc_df_list <- list()
for (m_label in names(roc_models)) {
  p_vec <- roc_models[[m_label]]
  if (!all(is.na(p_vec))) {
    roc_obj <- pROC::roc(response = y_true, predictor = p_vec, quiet = TRUE)
    auc_val <- as.numeric(pROC::auc(roc_obj))
    roc_df_list[[m_label]] <- data.frame(
      FPR = 1 - roc_obj$specificities,
      TPR = roc_obj$sensitivities,
      Model = paste0(m_label, " (AUC = ", round(auc_val, 3), ")")
    )
  }
}

roc_df_all <- bind_rows(roc_df_list)

# ggplot ROC curves
p_roc <- ggplot(roc_df_all, aes(x = FPR, y = TPR, color = Model)) +
  geom_line(linewidth = 1.2) +
  geom_abline(linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c(
    "#377eb8", "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#e6ab02"
  )) +
  theme_minimal(base_size = 13) +
  labs(
    title = "ROC Curves: SDM Variants & Ensemble vs Cull Dive COTS Presence (2025)",
    subtitle = paste0("n = ", nrow(cull_val_df), " independent cull dives from top 30 reefs"),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    color = "Model Variant"
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = c(0.68, 0.28),
    legend.background = element_rect(fill = "white", color = "grey80")
  ) +
  coord_equal()

print(p_roc)

# --- Comprehensive Metric Table on Cull Dives ---
cull_metrics_list <- list()

model_cols <- list(
  "Ensemble SDM (Weighted)"              = "ensemble",
  "Cull Dive XGBoost Reef Guide (17-pred)"= "cull_rg",
  "Cull Dive XGBoost Clean (12-pred)"     = "cull_clean",
  "Manta XGBoost Clean (12-pred)"        = "manta_clean",
  "Manta XGBoost Reef Guide (17-pred)"   = "manta_rg",
  "MaxEnt Clean (12-pred, 10k bg)"       = "maxent_clean",
  "MaxEnt Reef Guide (17-pred, 10k bg)"  = "maxent_rg"
)

for (m_name in names(model_cols)) {
  col_name <- model_cols[[m_name]]
  p_vec    <- cull_val_df[[col_name]]
  if (!all(is.na(p_vec))) {
    c_thresh <- median(p_vec, na.rm = TRUE)
    cull_metrics_list[[m_name]] <- compute_detailed_metrics(
      truth_vec = (cull_val_df$CPUE >= 0.02),
      prob_vec  = p_vec,
      model_name = m_name,
      dataset_name = "Independent Cull Validation",
      threshold = c_thresh
    )
  }
}

cull_perf_table <- bind_rows(cull_metrics_list)
print(kable(cull_perf_table, caption = "Independent Cull Validation Performance Metrics across All Base Models & Ensemble (Threshold = Median Probability)"))
cat("\n\n")

# --- Optimal thresholds (Youden's J) ---
roc_cull_rg <- pROC::roc(response = y_true, predictor = pull(cull_val_df, cull_rg), quiet = TRUE)
opt_cull_rg <- coords(roc_cull_rg, "best", best.method = "youden", ret = c("threshold", "sensitivity", "specificity"))

cat("Optimal Threshold (Youden's J) for Cull Dive XGBoost Reef Guide Model:", round(opt_cull_rg$threshold, 4),
    "  Sensitivity:", round(opt_cull_rg$sensitivity, 3),
    "  Specificity:", round(opt_cull_rg$specificity, 3), "\n\n")

# --- Confusion Matrices at P88 and P95 thresholds ---
thresh_rg_p88 <- quantile(cull_val_df$cull_rg, 0.88, na.rm = TRUE)
thresh_rg_p95 <- quantile(cull_val_df$cull_rg, 0.95, na.rm = TRUE)

plot_cull_cm <- function(prob_vec, obs_vec, threshold, title) {
  cm_df <- data.frame(
    Predicted = factor(ifelse(prob_vec >= threshold, "COTS Predicted", "No COTS Predicted"), 
                       levels = c("COTS Predicted", "No COTS Predicted")),
    Observed  = factor(ifelse(obs_vec, "COTS Present", "COTS Absent"), 
                       levels = c("COTS Present", "COTS Absent"))
  )
  cm_counts <- cm_df %>%
    count(Predicted, Observed) %>%
    group_by() %>%
    mutate(pct = n / sum(n) * 100)
  
  ggplot(cm_counts, aes(x = Predicted, y = Observed, fill = n)) +
    geom_tile(color = "white", linewidth = 2) +
    geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)")), size = 5, fontface = "bold") +
    scale_fill_gradient(low = "#f0f4ff", high = "#3366cc") +
    theme_minimal(base_size = 13) +
    labs(title = title, x = "SDM Prediction", y = "Observed", fill = "Count") +
    theme(plot.title = element_text(face = "bold", hjust = 0.5), panel.grid = element_blank())
}

p_cm1 <- plot_cull_cm(cull_val_df$cull_rg, cull_val_df$observed_presence == 1, 
                       thresh_rg_p88, paste0("Cull Dive ReefGuide @ P88 (", round(thresh_rg_p88, 4), ")"))
p_cm2 <- plot_cull_cm(cull_val_df$cull_rg, cull_val_df$observed_presence == 1, 
                       thresh_rg_p95, paste0("Cull Dive ReefGuide @ P95 (", round(thresh_rg_p95, 4), ")"))
p_cm3 <- plot_cull_cm(cull_val_df$cull_rg, cull_val_df$observed_presence == 1, 
                       opt_cull_rg$threshold, paste0("Cull Dive ReefGuide @ Optimal (", round(opt_cull_rg$threshold, 4), ")"))

gridExtra::grid.arrange(p_cm1, p_cm2, p_cm3, ncol = 3)
```

## 3. Step B: Tow-Path Simulation and Evaluation

Simulating whether a specific tow path would have been suggested by the 2025 SDM model. We evaluate hit-rates strictly against tows conducted from Jan 1 2025 onwards, however, we use all historic tow tracks to delineate the operational boundaries.

```{r step_b_tow_regions, cache=TRUE}

# 2. Define Regions (100m Buffered Tows & 200m Grids)
tow_regions_list <- list()
grid_regions_list <- list()

for (r_name in unique(manta_eval_all$ReefName)) {
  m_sub <- manta_eval_all %>% filter(ReefName == r_name)
  if (nrow(m_sub) == 0) next
  
  bb <- st_bbox(m_sub)
  gr_500 <- st_make_grid(bb, cellsize = 500, square = TRUE) %>% st_as_sf()
  
  # Tow Regions (100m buffer, unioned, then intersected with 500m grid)
  tr <- st_buffer(m_sub, dist = 100) %>% 
    st_union() %>% 
    st_intersection(gr_500) %>% 
    st_cast("POLYGON") %>% 
    st_as_sf() %>%
    mutate(ReefName = r_name, RegionType = "Tow_100m_Seg500m", RegionID = paste0(r_name, "_T_", row_number())) %>%
    rename(geometry = x)
  tow_regions_list[[r_name]] <- tr
  
  # Grid Regions (200m x 200m)
  gr <- st_make_grid(bb, cellsize = 200, square = TRUE) %>% 
    st_as_sf() %>%
    mutate(ReefName = r_name, RegionType = "Grid_200m", RegionID = paste0(r_name, "_G_", row_number())) %>%
    rename(geometry = x)
  
  # Keep only grid cells that intersect with manta tows to avoid evaluating empty ocean
  gr <- gr[lengths(st_intersects(gr, m_sub)) > 0, ]
  
  grid_regions_list[[r_name]] <- gr
}

tow_regions <- bind_rows(tow_regions_list)
grid_regions <- bind_rows(grid_regions_list)
all_regions <- bind_rows(tow_regions, grid_regions)

# 3. Extract SDM Stats for Regions
extract_poly_stats <- function(r, polys) {
  v <- terra::vect(polys)
  ex <- terra::extract(r, v)
  val_col <- setdiff(names(ex), "ID")[1]
  
  ex %>%
    rename(poly_id = ID, value = !!val_col) %>%
    group_by(poly_id) %>%
    summarise(
      reg_p88    = as.numeric(quantile(value, probs = 0.88, na.rm = TRUE, names = FALSE)),
      reg_p95    = as.numeric(quantile(value, probs = 0.95, na.rm = TRUE, names = FALSE)),
      .groups    = "drop"
    )
}

all_regions$poly_id <- seq_len(nrow(all_regions))
region_stats <- extract_poly_stats(prob_year2025, all_regions)

all_regions <- all_regions %>% left_join(region_stats, by = "poly_id")
tow_regions <- all_regions %>% filter(RegionType == "Tow_100m_Seg500m")
grid_regions <- all_regions %>% filter(RegionType == "Grid_200m")

# 4. Map Tows to Regions and Apply Thresholds
# using st_join(largest = TRUE) ensures a tow is assigned to only one region
m_tow <- st_join(manta_eval_all, tow_regions %>% select(RegionID_Tow = RegionID, reg_p88_Tow = reg_p88, reg_p95_Tow = reg_p95), largest = TRUE)
m_grid <- st_join(manta_eval_all, grid_regions %>% select(RegionID_Grid = RegionID, reg_p88_Grid = reg_p88, reg_p95_Grid = reg_p95), largest = TRUE)

manta_eval <- manta_eval_all %>%
  mutate(
    reg_p88_Tow = m_tow$reg_p88_Tow,
    reg_p95_Tow = m_tow$reg_p95_Tow,
    reg_p88_Grid = m_grid$reg_p88_Grid,
    reg_p95_Grid = m_grid$reg_p95_Grid
  ) %>%
  mutate(
    sug_Tow_P88 = !is.na(reg_p88_Tow) & reg_p88_Tow >= thresh_2025_p88,
    sug_Tow_P95 = !is.na(reg_p95_Tow) & reg_p95_Tow >= thresh_2025_p95,
    sug_Grid_P88 = !is.na(reg_p88_Grid) & reg_p88_Grid >= thresh_2025_p88,
    sug_Grid_P95 = !is.na(reg_p95_Grid) & reg_p95_Grid >= thresh_2025_p95
  )

# Function to compute hit rates by reef for a specific scenario
calc_hit_rates_reef <- function(df, scenario_col, scenario_name) {
  overall <- df %>%
    summarise(
      ReefName = "Overall",
      Scenario = scenario_name,
      Total_Tows = n(),
      Tows_Suggested = sum(!!sym(scenario_col), na.rm=TRUE),
      Hit_Rate = sum(!!sym(scenario_col) & presence_any, na.rm=TRUE) / Tows_Suggested,
      Tows_Avoided = sum(! (!!sym(scenario_col)), na.rm=TRUE),
      Miss_Rate = sum(! (!!sym(scenario_col)) & presence_any, na.rm=TRUE) / Tows_Avoided
    )
  by_reef <- df %>%
    group_by(ReefName) %>%
    summarise(
      Scenario = scenario_name,
      Total_Tows = n(),
      Tows_Suggested = sum(!!sym(scenario_col), na.rm=TRUE),
      Hit_Rate = sum(!!sym(scenario_col) & presence_any, na.rm=TRUE) / Tows_Suggested,
      Tows_Avoided = sum(! (!!sym(scenario_col)), na.rm=TRUE),
      Miss_Rate = sum(! (!!sym(scenario_col)) & presence_any, na.rm=TRUE) / Tows_Avoided
    )
  bind_rows(by_reef, overall)
}

manta_eval_2025 <- manta_eval %>% filter(as.Date(SurveyTime) >= as.Date("2025-01-01"))

hit_table <- bind_rows(
  calc_hit_rates_reef(manta_eval_2025 %>% filter(ReefName %in% reef_select), "sug_Tow_P88", "Seg Tow 100m + P88"),
  calc_hit_rates_reef(manta_eval_2025 %>% filter(ReefName %in% reef_select), "sug_Tow_P95", "Seg Tow 100m + P95")
) %>% arrange(ReefName, desc(Scenario))

kable(hit_table, digits = 3, caption = "Manta Tow Simulation: Operational Zonal Validation (Selected 10 Reefs)")
```

## 4. Step C: Cull Dive Financial Savings

Evaluating cull dives strictly from Jan 1 2025 onwards using the 2025 P88 and P95 thresholds on Segmented Tow Regions. We quantify avoided diving hours for dives that resulted in 0.00 CPUE (Ghost Culls) and <0.02 CPUE (Non-Problematic), and calculate financial savings ($950/hr).

```{r cull_financial_savings}
# --- Step 1: Build cull geometry using polygons where available, points otherwise ---
# Join cull data to polygon geometries by CullSiteName
cull_eval_raw <- cull_raw %>% filter(ReefName %in% reef_select_eval)
cull_eval_raw$row_id <- seq_len(nrow(cull_eval_raw))

# Match cull site names to polygon geometries
poly_lookup <- cull_site_polys %>% select(CullSiteName)

matched_idx <- cull_eval_raw$CullSiteName %in% poly_lookup$CullSiteName
cat("Cull site polygon match rate:", round(mean(matched_idx) * 100, 1), "%\n")
cat("  Matched:", sum(matched_idx), "  Unmatched:", sum(!matched_idx), "\n")

# For matched: use polygon + 50m buffer
if (sum(matched_idx) > 0) {
  # Look up the polygon for each matched cull record
  match_positions <- match(cull_eval_raw$CullSiteName[matched_idx], poly_lookup$CullSiteName)
  matched_geoms <- st_buffer(st_geometry(poly_lookup)[match_positions], dist = 50)
  cull_matched_sf2 <- st_sf(cull_eval_raw[matched_idx, ], geometry = matched_geoms)
  cull_matched_sf2$match_type <- "Polygon"
} else {
  cull_matched_sf2 <- NULL
}

# For unmatched: use point + 100m buffer
if (sum(!matched_idx) > 0) {
  cull_unmatched_pts <- st_as_sf(cull_eval_raw[!matched_idx, ], 
    coords = c("Longitude", "Latitude"), crs = 4326) %>%
    st_transform(crs = crs(prob_year2025))
  cull_unmatched_buf <- st_buffer(cull_unmatched_pts, dist = 100)
  cull_unmatched_buf$match_type <- "Point (100m buffer)"
} else {
  cull_unmatched_buf <- NULL
}

# Combine into single sf
cull_sf_combined <- bind_rows(cull_matched_sf2, cull_unmatched_buf) %>%
  arrange(row_id)

cat("Combined cull geometries:", nrow(cull_sf_combined), "rows\n")

# --- Step 2: Spatial join to tow/grid regions ---
c_tow  <- st_join(cull_sf_combined, tow_regions  %>% select(RegionID_Tow  = RegionID, reg_p88_Tow  = reg_p88, reg_p95_Tow  = reg_p95),  largest = TRUE)
c_grid <- st_join(cull_sf_combined, grid_regions %>% select(RegionID_Grid = RegionID, reg_p88_Grid = reg_p88, reg_p95_Grid = reg_p95), largest = TRUE)

# --- Step 3: Sector join (using centroids for sector assignment) ---
sectors_sf <- st_read("data/SectorShapefile/ltms.shp", quiet = TRUE) %>%
  st_transform(crs = crs(prob_year2025))
cull_centroids <- st_centroid(cull_sf_combined)
c_sector <- st_join(cull_centroids, sectors_sf %>% select(SECTOR_NAM), join = st_nearest_feature)

# --- Step 4: For unmatched tow-region NAs, extract MEAN raster value across the cull polygon/buffer ---
na_tow_idx <- is.na(c_tow$reg_p88_Tow)
cat("Tow region NA count:", sum(na_tow_idx), "of", nrow(c_tow), "\n")

# Extract mean raster value for each cull site polygon/buffer that didn't match a tow region
if (sum(na_tow_idx) > 0) {
  na_polys <- cull_sf_combined[na_tow_idx, ]
  na_extract <- terra::extract(prob_year2025, terra::vect(na_polys), fun = "mean", na.rm = TRUE)
  val_col <- setdiff(names(na_extract), "ID")[1]
  inferred_vals <- na_extract[[val_col]]
} else {
  inferred_vals <- numeric(0)
}

# --- Step 5: Build cull_eval with combined logic ---
cull_eval <- cull_sf_combined %>%
  st_drop_geometry() %>%
  mutate(
    reg_p88_Tow  = c_tow$reg_p88_Tow,
    reg_p95_Tow  = c_tow$reg_p95_Tow,
    reg_p88_Grid = c_grid$reg_p88_Grid,
    reg_p95_Grid = c_grid$reg_p95_Grid,
    SECTOR_NAM   = c_sector$SECTOR_NAM
  )

# For NAs: fill in with inferred raster mean value
cull_eval$inferred_mean <- NA_real_
cull_eval$inferred_mean[na_tow_idx] <- inferred_vals

cull_eval <- cull_eval %>%
  mutate(
    # Region-matched classification
    avoid_Tow_P88 = case_when(
      !is.na(reg_p88_Tow) ~ reg_p88_Tow < thresh_2025_p88,
      TRUE                 ~ inferred_mean < thresh_2025_p88
    ),
    avoid_Tow_P95 = case_when(
      !is.na(reg_p95_Tow) ~ reg_p95_Tow < thresh_2025_p95,
      TRUE                 ~ inferred_mean < thresh_2025_p95
    ),
    avoid_Grid_P88 = case_when(
      !is.na(reg_p88_Grid) ~ reg_p88_Grid < thresh_2025_p88,
      TRUE                  ~ inferred_mean < thresh_2025_p88
    ),
    avoid_Grid_P95 = case_when(
      !is.na(reg_p95_Grid) ~ reg_p95_Grid < thresh_2025_p95,
      TRUE                  ~ inferred_mean < thresh_2025_p95
    ),
    # Transparency flag
    classification_type = case_when(
      !is.na(reg_p88_Tow) ~ "Region Matched",
      !is.na(inferred_mean) ~ "Inferred",
      TRUE ~ "Unknown"
    )
  )

cat("\nClassification type breakdown:\n")
print(table(cull_eval$classification_type))

diver_hour_cost <- 950

calc_savings_reef <- function(df, avoid_col, scenario_name) {
  overall <- df %>%
    summarise(
      ReefName = "Overall",
      Scenario = scenario_name,
      Ghost_Avoided = sum(is_ghost & !!sym(avoid_col)),
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Avoided = sum(is_non_problematic & !!sym(avoid_col)),
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      Prob_Missed = sum(!is_non_problematic & !!sym(avoid_col)),
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0),
      Hours_Missed = sum(Bottomtime[!is_non_problematic & !!sym(avoid_col)]) / 60
    ) %>%
    mutate(
      CPUE_Missed = ifelse(Hours_Missed > 0, COTS_Missed / (Hours_Missed * 60), 0)
    )
  by_reef <- df %>%
    group_by(ReefName) %>%
    summarise(
      Scenario = scenario_name,
      Ghost_Avoided = sum(is_ghost & !!sym(avoid_col)),
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Avoided = sum(is_non_problematic & !!sym(avoid_col)),
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      Prob_Missed = sum(!is_non_problematic & !!sym(avoid_col)),
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0),
      Hours_Missed = sum(Bottomtime[!is_non_problematic & !!sym(avoid_col)]) / 60
    ) %>%
    mutate(
      CPUE_Missed = ifelse(Hours_Missed > 0, COTS_Missed / (Hours_Missed * 60), 0)
    )
  bind_rows(by_reef, overall)
}

cull_eval_2025 <- cull_eval %>% 
  filter(as.Date(SurveyDate) >= as.Date("2025-01-01")) %>%
  mutate(
    SECTOR_NAM = as.character(SECTOR_NAM),
    ManagementZone = case_when(
      SECTOR_NAM %in% c("Cape Grenville", "Princess Charlotte Bay") ~ "Far Northern",
      SECTOR_NAM %in% c("Cooktown / Lizard Island", "Cairns", "Innisfail") ~ "Northern",
      SECTOR_NAM %in% c("Townsville", "Cape Upstart", "Whitsunday") ~ "Central",
      SECTOR_NAM %in% c("Pompey", "Swain", "Capricorn Bunker") ~ "Southern",
      TRUE ~ "Unknown"
    )
  )

# Only show reef-level summary for the 10 selected reefs to avoid massive tables
savings_table <- bind_rows(
  calc_savings_reef(cull_eval_2025 %>% filter(ReefName %in% reef_select), "avoid_Tow_P88", "Seg Tow 100m + P88"),
  calc_savings_reef(cull_eval_2025 %>% filter(ReefName %in% reef_select), "avoid_Tow_P95", "Seg Tow 100m + P95")
) %>% arrange(ReefName, desc(Scenario))

kable(savings_table, digits = 2, format.args = list(big.mark = ","), caption = "Financial Savings and Risk Summary (Selected 10 Reefs)")

# Management Zone & Sector Summary (All 128 Reefs)
calc_savings_zone <- function(df, avoid_col, scenario_name) {
  df %>%
    group_by(ManagementZone, SECTOR_NAM) %>%
    summarise(
      Scenario = scenario_name,
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0),
      Hours_Missed = sum(Bottomtime[!is_non_problematic & !!sym(avoid_col)]) / 60,
      .groups = "drop"
    )
}

savings_zone_table <- bind_rows(
  calc_savings_zone(cull_eval_2025, "avoid_Tow_P88", "Seg Tow 100m + P88"),
  calc_savings_zone(cull_eval_2025, "avoid_Tow_P95", "Seg Tow 100m + P95")
) %>% 
  arrange(factor(ManagementZone, levels = c("Far Northern", "Northern", "Central", "Southern", "Unknown")), SECTOR_NAM, desc(Scenario))

kable(savings_zone_table, digits = 2, format.args = list(big.mark = ","), caption = "Financial Savings and Risk Summary by Management Zone & Sector (Top 30 Reefs, 2025)")

# --- Consolidated Scenario Summary ---
scenario_summary <- function(manta_df, cull_df, tow_col, avoid_col, label) {
  # Manta tow summary
  m <- manta_df %>%
    summarise(
      Total_Tows = n(),
      Tows_Suggested = sum(!!sym(tow_col), na.rm=TRUE),
      Tows_Avoided = Total_Tows - Tows_Suggested,
      Hit_Rate = sum(!!sym(tow_col) & presence_any, na.rm=TRUE) / max(Tows_Suggested, 1),
      Miss_Rate = sum(!(!!sym(tow_col)) & presence_any, na.rm=TRUE) / max(Tows_Avoided, 1)
    )
  # Cull dive summary
  c <- cull_df %>%
    summarise(
      Total_Dives = n(),
      Dives_Avoided = sum(!!sym(avoid_col), na.rm=TRUE),
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0)
    )
  tibble(
    Scenario = label,
    Total_Tows = m$Total_Tows, Tows_Suggested = m$Tows_Suggested, Tows_Avoided = m$Tows_Avoided,
    Tow_Hit_Rate = m$Hit_Rate, Tow_Miss_Rate = m$Miss_Rate,
    Total_Dives = c$Total_Dives, Dives_Avoided = c$Dives_Avoided,
    Total_Savings = c$Total_Savings, COTS_Missed = c$COTS_Missed, Pct_COTS_Missed = c$Pct_COTS_Missed
  )
}

consolidated <- bind_rows(
  scenario_summary(manta_eval_2025, cull_eval_2025, "sug_Tow_P88", "avoid_Tow_P88", "P88 (Aggressive)"),
  scenario_summary(manta_eval_2025, cull_eval_2025, "sug_Tow_P95", "avoid_Tow_P95", "P95 (Conservative)")
)

kable(consolidated, digits = 2, format.args = list(big.mark = ","), 
      caption = "Consolidated Scenario Comparison: Manta Tow + Cull Dive (All Top 30 Reefs, 2025)")
```

### Confusion Matrices

```{r confusion_matrices, fig.width=12, fig.height=5}
# Function to build and plot a confusion matrix
plot_confusion_matrix <- function(predicted_suggested, observed_presence, title) {
  cm_df <- data.frame(
    Predicted = factor(ifelse(predicted_suggested, "Suggested", "Avoided"), levels = c("Suggested", "Avoided")),
    Observed  = factor(ifelse(observed_presence, "COTS Present", "COTS Absent"), levels = c("COTS Present", "COTS Absent"))
  )
  cm_counts <- cm_df %>%
    count(Predicted, Observed) %>%
    group_by() %>%
    mutate(pct = n / sum(n) * 100)
  
  ggplot(cm_counts, aes(x = Predicted, y = Observed, fill = n)) +
    geom_tile(color = "white", linewidth = 2) +
    geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)") ), size = 5, fontface = "bold") +
    scale_fill_gradient(low = "#f0f4ff", high = "#3366cc") +
    theme_minimal(base_size = 14) +
    labs(title = title, x = "SDM Prediction", y = "Observed", fill = "Count") +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      panel.grid = element_blank()
    )
}

# Manta Tow Confusion Matrices
p1 <- plot_confusion_matrix(
  manta_eval_2025$sug_Tow_P88, manta_eval_2025$presence_any,
  "Manta Tow — P88 (Aggressive)"
)
p2 <- plot_confusion_matrix(
  manta_eval_2025$sug_Tow_P95, manta_eval_2025$presence_any,
  "Manta Tow — P95 (Conservative)"
)

gridExtra::grid.arrange(p1, p2, ncol = 2)

# Cull Dive Confusion Matrices (is_non_problematic as "COTS Absent" proxy)
p3 <- plot_confusion_matrix(
  !cull_eval_2025$avoid_Tow_P88, !cull_eval_2025$is_non_problematic,
  "Cull Dive — P88 (Aggressive)"
)
p4 <- plot_confusion_matrix(
  !cull_eval_2025$avoid_Tow_P95, !cull_eval_2025$is_non_problematic,
  "Cull Dive — P95 (Conservative)"
)

gridExtra::grid.arrange(p3, p4, ncol = 2)
```

## 5. Step D: Visual Mapping

Below are maps for each reef showing the underlying 2025 SDM probability. The operational units displayed are the **100m Buffered Tow Regions**, categorized by the **95th Percentile (P95) Threshold** (Red = Suggested, Grey = Avoided). Individual manta tows and cull dive locations are also plotted to show what falls inside the suggested operational areas.

```{r visual_mapping, results='asis', fig.width=10, fig.height=7}
plot_reef_map <- function(r, tow_sf, dive_sf, region_sf, reef_name, p_min, p_max) {
  # Bounding box around features
  bb_t <- if(nrow(tow_sf) > 0) st_bbox(tow_sf) else NULL
  bb_d <- if(nrow(dive_sf) > 0) st_bbox(dive_sf) else NULL
  
  xmin <- min(c(bb_t["xmin"], bb_d["xmin"]), na.rm=TRUE)
  ymin <- min(c(bb_t["ymin"], bb_d["ymin"]), na.rm=TRUE)
  xmax <- max(c(bb_t["xmax"], bb_d["xmax"]), na.rm=TRUE)
  ymax <- max(c(bb_t["ymax"], bb_d["ymax"]), na.rm=TRUE)
  
  pad <- 2000 # 2km padding
  ext_map <- terra::ext(xmin - pad, xmax + pad, ymin - pad, ymax + pad)
  
  rc <- terra::crop(r, ext_map)
  
  # Scale aggregate factor to keep cell count manageable (~10,000 pixels) for fast plotting
  fact <- ceiling(sqrt(terra::ncell(rc) / 10000))
  if(fact > 1) {
    rc <- terra::aggregate(rc, fact = fact, fun = "mean", na.rm = TRUE)
  }
  
  rdf <- as.data.frame(rc, xy = TRUE, na.rm = TRUE)
  names(rdf)[3] <- "prob"
  
  # Determine if region is suggested based on P95 threshold
  region_sf <- region_sf %>%
    mutate(is_suggested = !is.na(reg_p95) & reg_p95 >= thresh_2025_p95)
  
  gg <- ggplot() +
    geom_raster(data = rdf, aes(x = x, y = y, fill = prob)) +
    # Plot Operational Zones
    geom_sf(data = region_sf, aes(color = is_suggested), fill = NA, linewidth = 1.2, linetype = "dashed") +
    scale_color_manual(values = c("TRUE" = "red", "FALSE" = "grey70")) +
    # Plot Tows and Dives
    geom_sf(data = tow_sf, color = "black", linewidth = 0.5) +
    geom_sf(data = dive_sf, shape = 21, fill = "white", color = "black", size = 2) +
    scale_fill_viridis_c(
      option = "C", 
      na.value = NA, 
      limits = c(p_min, p_max), 
      oob = scales::squish
    ) +
    theme_minimal() +
    labs(
      title = paste("Operational Zonal Map:", reef_name),
      subtitle = paste0("Red Dashed = Suggested 100m Zone (P95), Grey Dashed = Avoided Zone\nScale clipped to ", p_min, "-", p_max),
      fill  = "2025 SDM Prob",
      color = "Suggested Zone"
    )
  return(gg)
}

# Generate plots for each reef
cull_sf_2025 <- cull_sf %>% filter(as.Date(SurveyDate) >= as.Date("2025-01-01"))

for(r_name in reef_select) {
  t_sf <- manta_eval_2025 %>% filter(ReefName == r_name)
  d_sf <- cull_sf_2025 %>% filter(ReefName == r_name)
  r_sf <- tow_regions %>% filter(ReefName == r_name)
  
  if(nrow(t_sf) > 0 || nrow(d_sf) > 0) {
    message(paste("Processing map for:", r_name))
    cat("\n### ", r_name, "\n")
    p <- plot_reef_map(prob_year2025, t_sf, d_sf, r_sf, r_name, p_min_2025, p_max_2025)
    print(p)
    cat("\n")
  }
}
```
