---
title: "COTS Fine-Scale SDM: Comprehensive Validation Report"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
    fig-width: 10
    fig-height: 8
  pdf:
    toc: true
execute:
  warning: false
  message: false
---

## Overview

This report validates the XGBoost COTS Fine-Scale Species Distribution Model using historical Manta Tow and Cull data. We compare both the **2025 Year-Specific** and **Year-Agnostic** models.

The validation involves:
1. **Grid-Based Historical Validation (100m)**: General statistics (ROC, AUC, Confusion Matrix) evaluating the model against historical COTS/scar presence.
2. **Tow-Path Simulation**: Operational simulation showing which manta tows would be suggested based on the model.
3. **Cull Dive Financial Savings**: Quantification of dive hours and dollars saved by avoiding ineffective culls (0.00 or <0.02 CPUE).
4. **Reef Maps**: Visual mapping of predictions vs. operations.

```{r setup}
library(terra)
library(sf)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
library(tidyr)
library(knitr)
library(stringr)
library(pROC)

# --- Configuration ---
predict_agnostic_path <- "outputs/COTS_prob_0.02_cpue_agnostic.tif"
predict_year2025_path <- "outputs/COTS_prob_0.02_cpue_year2025.tif"  # native 30m raster
manta_csv <- "data/COTS Program  Manta Tow Data-2026-02-04.csv"
cull_xlsx <- "data/250929_COTS-Manta-Cull-RHIS-Data-Matthews-and-Schlawinsky.xlsx"
cull_polys_gpkg <- "data/cull_site_polygons.gpkg"

out_dir <- "validation_outputs"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Selected Reefs
reef_select <- c(
  "Tongue Reef (16-026)",
  "Batt Reef (16-029)",
  "Arlington Reef (16-064)",
  "(Big) Broadhurst Reef (No 1) (18-100a)",
  "Britomart Reef (18-024)",
  "Sudbury Reef (No 1) (17-001a)",
  "Trunk Reef (18-027)",
  "Bramble Reef (18-029)",
  "Davies Reef (18-096)",
  "Heron Reef (23-052a)"
)
```

## 1. Load and Prepare Data

```{r load_data}
# Load Model Predictions
prob_agnostic <- if(file.exists(predict_agnostic_path)) terra::rast(predict_agnostic_path) else NULL
prob_year2025 <- if(file.exists(predict_year2025_path)) terra::rast(predict_year2025_path) else NULL

if(is.null(prob_year2025) || is.null(prob_agnostic)) stop("Missing prediction rasters.")

# Load Manta Tow Data
manta_raw <- read_csv(manta_csv, show_col_types = FALSE) %>%
  filter(
    !is.na(StartLongitude), !is.na(StartLatitude), !is.na(EndLongitude), !is.na(EndLatitude),
    StartLongitude != 0, StartLatitude != 0, EndLongitude != 0, EndLatitude != 0
  ) %>%
  mutate(
    cots_observed = CrownOfThornsStarfishCount > 0,
    scars_observed = !is.na(ScarsCount) & ScarsCount > 0,
    presence_any = cots_observed | scars_observed,
    ReefName = as.character(ReefName)
  )

# Convert Manta Tows to spatial lines
make_tow_lines <- function(df) {
  lines <- lapply(seq_len(nrow(df)), function(i) {
    p1 <- c(df$StartLongitude[i], df$StartLatitude[i])
    p2 <- c(df$EndLongitude[i], df$EndLatitude[i])
    st_linestring(rbind(p1, p2))
  })
  st_sf(df, geometry = st_sfc(lines, crs = 4326))
}

manta_sf <- make_tow_lines(manta_raw)
manta_sf <- st_transform(manta_sf, crs = crs(prob_year2025))
manta_selected <- manta_sf %>% filter(ReefName %in% reef_select)

# Load Cull Data
cull_raw <- read_excel(cull_xlsx, sheet = 4) %>%
  filter(!is.na(Longitude), !is.na(Latitude), !is.na(Bottomtime), Bottomtime > 0, Longitude != 0, Latitude != 0) %>%
  filter(as.Date(SurveyDate) > as.Date("2018-11-30")) %>%
  mutate(
    Total = Cohort1 + Cohort2 + Cohort3 + Cohort4,
    CPUE = Total / Bottomtime,
    is_ghost = (Total == 0),
    is_non_problematic = (CPUE < 0.02),
    ReefName = as.character(ReefName)
  )

# Top 30 reefs with most culling hours in 2025
reef_select_eval <- cull_raw %>% 
  filter(as.Date(SurveyDate) >= as.Date("2025-01-01")) %>%
  group_by(ReefName) %>%
  summarise(Total_Hours = sum(Bottomtime, na.rm=TRUE) / 60) %>%
  arrange(desc(Total_Hours)) %>%
  slice_head(n = 30) %>%
  pull(ReefName)

# Manta tows for the selected 10 mapping reefs AND the 30 evaluated reefs
manta_selected <- manta_sf %>% filter(ReefName %in% reef_select)
manta_eval_all <- manta_sf %>% filter(ReefName %in% reef_select_eval)

# Cull dives for the selected 10 mapping reefs
cull_selected <- cull_raw %>% filter(ReefName %in% reef_select)
cull_sf <- st_as_sf(cull_selected, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = crs(prob_year2025))

# Cull dives for the 30 evaluated reefs (point geometry)
cull_sf_eval_all <- st_as_sf(cull_raw %>% filter(ReefName %in% reef_select_eval), coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(crs = crs(prob_year2025))

# Load cull site polygons and join to cull data
cull_site_polys <- st_read(cull_polys_gpkg, quiet = TRUE) %>%
  st_transform(crs = crs(prob_year2025))

cat("Loaded", nrow(cull_site_polys), "cull site polygons from gpkg\n")
cat("Matching to", n_distinct(cull_raw$CullSiteName), "unique cull site names...\n")
cat("Match rate:", round(mean(cull_raw$CullSiteName %in% cull_site_polys$CullSiteName) * 100, 1), "%\n")

# Calculate Default Threshold: Median of 2025 raster cropped to selected reefs
ext_val <- terra::ext(manta_selected) + 2000
crop_2025 <- terra::crop(prob_year2025, ext_val)
median_thresh_2025 <- as.numeric(terra::global(crop_2025, fun=function(x) median(x, na.rm=TRUE)))

crop_agnostic <- terra::crop(prob_agnostic, ext_val)
median_thresh_agnostic <- as.numeric(terra::global(crop_agnostic, fun=function(x) median(x, na.rm=TRUE)))

# Fixed mapping cutoffs
p_min_2025 <- 0.48
p_max_2025 <- 0.55

cat("Median Threshold (2025):", round(median_thresh_2025, 3), "\n")
cat("Median Threshold (Agnostic):", round(median_thresh_agnostic, 3), "\n")
cat("Mapping Range (2025):", p_min_2025, "to", p_max_2025, "\n")
```

## 2. Step A: Grid-Based Historical Validation (100m)

We rasterize historical manta tow observations onto a 100m grid and compare them to the SDM probability values (using both 2025 and Agnostic models). We calculate ROC curves, AUC, and evaluate thresholds using confusion matrices.

```{r step_a_grid, fig.width=12, fig.height=6}
if(nrow(manta_selected) > 0) {
  # crop_2025 is already 60m. Resample crop_agnostic to match it exactly.
  pred_60m_2025 <- crop_2025
  pred_60m_agn <- terra::resample(crop_agnostic, crop_2025, method = "bilinear")
  
  # Rasterize Manta Tow presence (1 if ever present, 0 if absent)
  manta_presence <- terra::rasterize(
    terra::vect(manta_selected), 
    pred_60m_2025, 
    field = "presence_any", 
    fun = "max", 
    background = NA
  )
  
  # Build evaluation dataframe
  val_df <- data.frame(
    observed_presence = terra::values(manta_presence)[,1],
    prob_2025 = terra::values(pred_60m_2025)[,1],
    prob_agnostic = terra::values(pred_60m_agn)[,1]
  ) %>% drop_na()
  
  # --- ROC and AUC ---
  roc_2025 <- roc(val_df$observed_presence, val_df$prob_2025, quiet=TRUE)
  roc_agn <- roc(val_df$observed_presence, val_df$prob_agnostic, quiet=TRUE)
  
  cat("AUC (2025 Model):", round(auc(roc_2025), 3), "\n")
  cat("AUC (Agnostic Model):", round(auc(roc_agn), 3), "\n\n")
  
  # Plot ROC
  par(mfrow=c(1,2))
  plot(roc_2025, main="ROC Curve (2025)", col="blue", lwd=2)
  plot(roc_agn, main="ROC Curve (Agnostic)", col="red", lwd=2)
  
  # --- Confusion Matrix at Median Threshold ---
  cat("Confusion Matrix for 2025 Model (Threshold = Median:", round(median_thresh_2025,3), ")\n")
  cm_2025 <- table(
    Predicted = factor(val_df$prob_2025 >= median_thresh_2025, levels=c(FALSE, TRUE)),
    Observed = factor(val_df$observed_presence == 1, levels=c(FALSE, TRUE))
  )
  print(cm_2025)
  cat("Accuracy:", sum(diag(cm_2025))/sum(cm_2025), "\n\n")
  
  cat("Confusion Matrix for Agnostic Model (Threshold = Median:", round(median_thresh_agnostic,3), ")\n")
  cm_agn <- table(
    Predicted = factor(val_df$prob_agnostic >= median_thresh_agnostic, levels=c(FALSE, TRUE)),
    Observed = factor(val_df$observed_presence == 1, levels=c(FALSE, TRUE))
  )
  print(cm_agn)
  cat("Accuracy:", sum(diag(cm_agn))/sum(cm_agn), "\n")
}
```

## 3. Step B: Tow-Path Simulation and Evaluation

Simulating whether a specific tow path would have been suggested by the 2025 SDM model. We evaluate hit-rates strictly against tows conducted from Jan 1 2025 onwards, however, we use all historic tow tracks to delineate the operational boundaries.

```{r step_b_tow_regions, cache=TRUE}
# 1. Calculate Global 2025 Raster Thresholds
thresh_2025_p88 <- as.numeric(terra::global(crop_2025, fun=function(x) quantile(x, probs=0.88, na.rm=TRUE)))
thresh_2025_p95 <- as.numeric(terra::global(crop_2025, fun=function(x) quantile(x, probs=0.95, na.rm=TRUE)))

cat("Global Thresholds for 2025 Validation:\n")
cat("- 88th Percentile:", round(thresh_2025_p88, 3), "\n")
cat("- 95th Percentile:", round(thresh_2025_p95, 3), "\n\n")

# 2. Define Regions (100m Buffered Tows & 200m Grids)
tow_regions_list <- list()
grid_regions_list <- list()

for (r_name in unique(manta_eval_all$ReefName)) {
  m_sub <- manta_eval_all %>% filter(ReefName == r_name)
  if (nrow(m_sub) == 0) next
  
  bb <- st_bbox(m_sub)
  gr_500 <- st_make_grid(bb, cellsize = 500, square = TRUE) %>% st_as_sf()
  
  # Tow Regions (100m buffer, unioned, then intersected with 500m grid)
  tr <- st_buffer(m_sub, dist = 100) %>% 
    st_union() %>% 
    st_intersection(gr_500) %>% 
    st_cast("POLYGON") %>% 
    st_as_sf() %>%
    mutate(ReefName = r_name, RegionType = "Tow_100m_Seg500m", RegionID = paste0(r_name, "_T_", row_number())) %>%
    rename(geometry = x)
  tow_regions_list[[r_name]] <- tr
  
  # Grid Regions (200m x 200m)
  gr <- st_make_grid(bb, cellsize = 200, square = TRUE) %>% 
    st_as_sf() %>%
    mutate(ReefName = r_name, RegionType = "Grid_200m", RegionID = paste0(r_name, "_G_", row_number())) %>%
    rename(geometry = x)
  
  # Keep only grid cells that intersect with manta tows to avoid evaluating empty ocean
  gr <- gr[lengths(st_intersects(gr, m_sub)) > 0, ]
  
  grid_regions_list[[r_name]] <- gr
}

tow_regions <- bind_rows(tow_regions_list)
grid_regions <- bind_rows(grid_regions_list)
all_regions <- bind_rows(tow_regions, grid_regions)

# 3. Extract SDM Stats for Regions
extract_poly_stats <- function(r, polys) {
  v <- terra::vect(polys)
  ex <- terra::extract(r, v)
  val_col <- setdiff(names(ex), "ID")[1]
  
  ex %>%
    rename(poly_id = ID, value = !!val_col) %>%
    group_by(poly_id) %>%
    summarise(
      reg_p88    = as.numeric(quantile(value, probs = 0.88, na.rm = TRUE, names = FALSE)),
      reg_p95    = as.numeric(quantile(value, probs = 0.95, na.rm = TRUE, names = FALSE)),
      .groups    = "drop"
    )
}

all_regions$poly_id <- seq_len(nrow(all_regions))
region_stats <- extract_poly_stats(prob_year2025, all_regions)

all_regions <- all_regions %>% left_join(region_stats, by = "poly_id")
tow_regions <- all_regions %>% filter(RegionType == "Tow_100m_Seg500m")
grid_regions <- all_regions %>% filter(RegionType == "Grid_200m")

# 4. Map Tows to Regions and Apply Thresholds
# using st_join(largest = TRUE) ensures a tow is assigned to only one region
m_tow <- st_join(manta_eval_all, tow_regions %>% select(RegionID_Tow = RegionID, reg_p88_Tow = reg_p88, reg_p95_Tow = reg_p95), largest = TRUE)
m_grid <- st_join(manta_eval_all, grid_regions %>% select(RegionID_Grid = RegionID, reg_p88_Grid = reg_p88, reg_p95_Grid = reg_p95), largest = TRUE)

manta_eval <- manta_eval_all %>%
  mutate(
    reg_p88_Tow = m_tow$reg_p88_Tow,
    reg_p95_Tow = m_tow$reg_p95_Tow,
    reg_p88_Grid = m_grid$reg_p88_Grid,
    reg_p95_Grid = m_grid$reg_p95_Grid
  ) %>%
  mutate(
    sug_Tow_P88 = !is.na(reg_p88_Tow) & reg_p88_Tow >= thresh_2025_p88,
    sug_Tow_P95 = !is.na(reg_p95_Tow) & reg_p95_Tow >= thresh_2025_p95,
    sug_Grid_P88 = !is.na(reg_p88_Grid) & reg_p88_Grid >= thresh_2025_p88,
    sug_Grid_P95 = !is.na(reg_p95_Grid) & reg_p95_Grid >= thresh_2025_p95
  )

# Function to compute hit rates by reef for a specific scenario
calc_hit_rates_reef <- function(df, scenario_col, scenario_name) {
  overall <- df %>%
    summarise(
      ReefName = "Overall",
      Scenario = scenario_name,
      Total_Tows = n(),
      Tows_Suggested = sum(!!sym(scenario_col), na.rm=TRUE),
      Hit_Rate = sum(!!sym(scenario_col) & presence_any, na.rm=TRUE) / Tows_Suggested,
      Tows_Avoided = sum(! (!!sym(scenario_col)), na.rm=TRUE),
      Miss_Rate = sum(! (!!sym(scenario_col)) & presence_any, na.rm=TRUE) / Tows_Avoided
    )
  by_reef <- df %>%
    group_by(ReefName) %>%
    summarise(
      Scenario = scenario_name,
      Total_Tows = n(),
      Tows_Suggested = sum(!!sym(scenario_col), na.rm=TRUE),
      Hit_Rate = sum(!!sym(scenario_col) & presence_any, na.rm=TRUE) / Tows_Suggested,
      Tows_Avoided = sum(! (!!sym(scenario_col)), na.rm=TRUE),
      Miss_Rate = sum(! (!!sym(scenario_col)) & presence_any, na.rm=TRUE) / Tows_Avoided
    )
  bind_rows(by_reef, overall)
}

manta_eval_2025 <- manta_eval %>% filter(as.Date(SurveyTime) >= as.Date("2025-01-01"))

hit_table <- bind_rows(
  calc_hit_rates_reef(manta_eval_2025 %>% filter(ReefName %in% reef_select), "sug_Tow_P88", "Seg Tow 100m + P88"),
  calc_hit_rates_reef(manta_eval_2025 %>% filter(ReefName %in% reef_select), "sug_Tow_P95", "Seg Tow 100m + P95")
) %>% arrange(ReefName, desc(Scenario))

kable(hit_table, digits = 3, caption = "Manta Tow Simulation: Operational Zonal Validation (Selected 10 Reefs)")
```

## 4. Step C: Cull Dive Financial Savings

Evaluating cull dives strictly from Jan 1 2025 onwards using the 2025 P88 and P95 thresholds on Segmented Tow Regions. We quantify avoided diving hours for dives that resulted in 0.00 CPUE (Ghost Culls) and <0.02 CPUE (Non-Problematic), and calculate financial savings ($950/hr).

```{r cull_financial_savings}
# --- Step 1: Build cull geometry using polygons where available, points otherwise ---
# Join cull data to polygon geometries by CullSiteName
cull_eval_raw <- cull_raw %>% filter(ReefName %in% reef_select_eval)
cull_eval_raw$row_id <- seq_len(nrow(cull_eval_raw))

# Match cull site names to polygon geometries
poly_lookup <- cull_site_polys %>% select(CullSiteName)

matched_idx <- cull_eval_raw$CullSiteName %in% poly_lookup$CullSiteName
cat("Cull site polygon match rate:", round(mean(matched_idx) * 100, 1), "%\n")
cat("  Matched:", sum(matched_idx), "  Unmatched:", sum(!matched_idx), "\n")

# For matched: use polygon + 50m buffer
if (sum(matched_idx) > 0) {
  # Look up the polygon for each matched cull record
  match_positions <- match(cull_eval_raw$CullSiteName[matched_idx], poly_lookup$CullSiteName)
  matched_geoms <- st_buffer(st_geometry(poly_lookup)[match_positions], dist = 50)
  cull_matched_sf2 <- st_sf(cull_eval_raw[matched_idx, ], geometry = matched_geoms)
  cull_matched_sf2$match_type <- "Polygon"
} else {
  cull_matched_sf2 <- NULL
}

# For unmatched: use point + 100m buffer
if (sum(!matched_idx) > 0) {
  cull_unmatched_pts <- st_as_sf(cull_eval_raw[!matched_idx, ], 
    coords = c("Longitude", "Latitude"), crs = 4326) %>%
    st_transform(crs = crs(prob_year2025))
  cull_unmatched_buf <- st_buffer(cull_unmatched_pts, dist = 100)
  cull_unmatched_buf$match_type <- "Point (100m buffer)"
} else {
  cull_unmatched_buf <- NULL
}

# Combine into single sf
cull_sf_combined <- bind_rows(cull_matched_sf2, cull_unmatched_buf) %>%
  arrange(row_id)

cat("Combined cull geometries:", nrow(cull_sf_combined), "rows\n")

# --- Step 2: Spatial join to tow/grid regions ---
c_tow  <- st_join(cull_sf_combined, tow_regions  %>% select(RegionID_Tow  = RegionID, reg_p88_Tow  = reg_p88, reg_p95_Tow  = reg_p95),  largest = TRUE)
c_grid <- st_join(cull_sf_combined, grid_regions %>% select(RegionID_Grid = RegionID, reg_p88_Grid = reg_p88, reg_p95_Grid = reg_p95), largest = TRUE)

# --- Step 3: Sector join (using centroids for sector assignment) ---
sectors_sf <- st_read("data/SectorShapefile/ltms.shp", quiet = TRUE) %>%
  st_transform(crs = crs(prob_year2025))
cull_centroids <- st_centroid(cull_sf_combined)
c_sector <- st_join(cull_centroids, sectors_sf %>% select(SECTOR_NAM), join = st_nearest_feature)

# --- Step 4: For unmatched tow-region NAs, extract MEAN raster value across the cull polygon/buffer ---
na_tow_idx <- is.na(c_tow$reg_p88_Tow)
cat("Tow region NA count:", sum(na_tow_idx), "of", nrow(c_tow), "\n")

# Extract mean raster value for each cull site polygon/buffer that didn't match a tow region
if (sum(na_tow_idx) > 0) {
  na_polys <- cull_sf_combined[na_tow_idx, ]
  na_extract <- terra::extract(prob_year2025, terra::vect(na_polys), fun = "mean", na.rm = TRUE)
  val_col <- setdiff(names(na_extract), "ID")[1]
  inferred_vals <- na_extract[[val_col]]
} else {
  inferred_vals <- numeric(0)
}

# --- Step 5: Build cull_eval with combined logic ---
cull_eval <- cull_sf_combined %>%
  st_drop_geometry() %>%
  mutate(
    reg_p88_Tow  = c_tow$reg_p88_Tow,
    reg_p95_Tow  = c_tow$reg_p95_Tow,
    reg_p88_Grid = c_grid$reg_p88_Grid,
    reg_p95_Grid = c_grid$reg_p95_Grid,
    SECTOR_NAM   = c_sector$SECTOR_NAM
  )

# For NAs: fill in with inferred raster mean value
cull_eval$inferred_mean <- NA_real_
cull_eval$inferred_mean[na_tow_idx] <- inferred_vals

cull_eval <- cull_eval %>%
  mutate(
    # Region-matched classification
    avoid_Tow_P88 = case_when(
      !is.na(reg_p88_Tow) ~ reg_p88_Tow < thresh_2025_p88,
      TRUE                 ~ inferred_mean < thresh_2025_p88
    ),
    avoid_Tow_P95 = case_when(
      !is.na(reg_p95_Tow) ~ reg_p95_Tow < thresh_2025_p95,
      TRUE                 ~ inferred_mean < thresh_2025_p95
    ),
    avoid_Grid_P88 = case_when(
      !is.na(reg_p88_Grid) ~ reg_p88_Grid < thresh_2025_p88,
      TRUE                  ~ inferred_mean < thresh_2025_p88
    ),
    avoid_Grid_P95 = case_when(
      !is.na(reg_p95_Grid) ~ reg_p95_Grid < thresh_2025_p95,
      TRUE                  ~ inferred_mean < thresh_2025_p95
    ),
    # Transparency flag
    classification_type = case_when(
      !is.na(reg_p88_Tow) ~ "Region Matched",
      !is.na(inferred_mean) ~ "Inferred",
      TRUE ~ "Unknown"
    )
  )

cat("\nClassification type breakdown:\n")
print(table(cull_eval$classification_type))

diver_hour_cost <- 950

calc_savings_reef <- function(df, avoid_col, scenario_name) {
  overall <- df %>%
    summarise(
      ReefName = "Overall",
      Scenario = scenario_name,
      Ghost_Avoided = sum(is_ghost & !!sym(avoid_col)),
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Avoided = sum(is_non_problematic & !!sym(avoid_col)),
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      Prob_Missed = sum(!is_non_problematic & !!sym(avoid_col)),
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0),
      Hours_Missed = sum(Bottomtime[!is_non_problematic & !!sym(avoid_col)]) / 60
    ) %>%
    mutate(
      CPUE_Missed = ifelse(Hours_Missed > 0, COTS_Missed / (Hours_Missed * 60), 0)
    )
  by_reef <- df %>%
    group_by(ReefName) %>%
    summarise(
      Scenario = scenario_name,
      Ghost_Avoided = sum(is_ghost & !!sym(avoid_col)),
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Avoided = sum(is_non_problematic & !!sym(avoid_col)),
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      Prob_Missed = sum(!is_non_problematic & !!sym(avoid_col)),
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0),
      Hours_Missed = sum(Bottomtime[!is_non_problematic & !!sym(avoid_col)]) / 60
    ) %>%
    mutate(
      CPUE_Missed = ifelse(Hours_Missed > 0, COTS_Missed / (Hours_Missed * 60), 0)
    )
  bind_rows(by_reef, overall)
}

cull_eval_2025 <- cull_eval %>% 
  filter(as.Date(SurveyDate) >= as.Date("2025-01-01")) %>%
  mutate(
    SECTOR_NAM = as.character(SECTOR_NAM),
    ManagementZone = case_when(
      SECTOR_NAM %in% c("Cape Grenville", "Princess Charlotte Bay") ~ "Far Northern",
      SECTOR_NAM %in% c("Cooktown / Lizard Island", "Cairns", "Innisfail") ~ "Northern",
      SECTOR_NAM %in% c("Townsville", "Cape Upstart", "Whitsunday") ~ "Central",
      SECTOR_NAM %in% c("Pompey", "Swain", "Capricorn Bunker") ~ "Southern",
      TRUE ~ "Unknown"
    )
  )

# Only show reef-level summary for the 10 selected reefs to avoid massive tables
savings_table <- bind_rows(
  calc_savings_reef(cull_eval_2025 %>% filter(ReefName %in% reef_select), "avoid_Tow_P88", "Seg Tow 100m + P88"),
  calc_savings_reef(cull_eval_2025 %>% filter(ReefName %in% reef_select), "avoid_Tow_P95", "Seg Tow 100m + P95")
) %>% arrange(ReefName, desc(Scenario))

kable(savings_table, digits = 2, format.args = list(big.mark = ","), caption = "Financial Savings and Risk Summary (Selected 10 Reefs)")

# Management Zone & Sector Summary (All 128 Reefs)
calc_savings_zone <- function(df, avoid_col, scenario_name) {
  df %>%
    group_by(ManagementZone, SECTOR_NAM) %>%
    summarise(
      Scenario = scenario_name,
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0),
      Hours_Missed = sum(Bottomtime[!is_non_problematic & !!sym(avoid_col)]) / 60,
      .groups = "drop"
    )
}

savings_zone_table <- bind_rows(
  calc_savings_zone(cull_eval_2025, "avoid_Tow_P88", "Seg Tow 100m + P88"),
  calc_savings_zone(cull_eval_2025, "avoid_Tow_P95", "Seg Tow 100m + P95")
) %>% 
  arrange(factor(ManagementZone, levels = c("Far Northern", "Northern", "Central", "Southern", "Unknown")), SECTOR_NAM, desc(Scenario))

kable(savings_zone_table, digits = 2, format.args = list(big.mark = ","), caption = "Financial Savings and Risk Summary by Management Zone & Sector (Top 30 Reefs, 2025)")

# --- Consolidated Scenario Summary ---
scenario_summary <- function(manta_df, cull_df, tow_col, avoid_col, label) {
  # Manta tow summary
  m <- manta_df %>%
    summarise(
      Total_Tows = n(),
      Tows_Suggested = sum(!!sym(tow_col), na.rm=TRUE),
      Tows_Avoided = Total_Tows - Tows_Suggested,
      Hit_Rate = sum(!!sym(tow_col) & presence_any, na.rm=TRUE) / max(Tows_Suggested, 1),
      Miss_Rate = sum(!(!!sym(tow_col)) & presence_any, na.rm=TRUE) / max(Tows_Avoided, 1)
    )
  # Cull dive summary
  c <- cull_df %>%
    summarise(
      Total_Dives = n(),
      Dives_Avoided = sum(!!sym(avoid_col), na.rm=TRUE),
      Ghost_Savings = sum(Bottomtime[is_ghost & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      NonProb_Savings = sum(Bottomtime[is_non_problematic & !!sym(avoid_col)]) / 60 * diver_hour_cost,
      Total_Savings = Ghost_Savings + NonProb_Savings,
      COTS_Missed = sum(Total[!is_non_problematic & !!sym(avoid_col)]),
      Pct_COTS_Missed = ifelse(sum(Total) > 0, COTS_Missed / sum(Total) * 100, 0)
    )
  tibble(
    Scenario = label,
    Total_Tows = m$Total_Tows, Tows_Suggested = m$Tows_Suggested, Tows_Avoided = m$Tows_Avoided,
    Tow_Hit_Rate = m$Hit_Rate, Tow_Miss_Rate = m$Miss_Rate,
    Total_Dives = c$Total_Dives, Dives_Avoided = c$Dives_Avoided,
    Total_Savings = c$Total_Savings, COTS_Missed = c$COTS_Missed, Pct_COTS_Missed = c$Pct_COTS_Missed
  )
}

consolidated <- bind_rows(
  scenario_summary(manta_eval_2025, cull_eval_2025, "sug_Tow_P88", "avoid_Tow_P88", "P88 (Aggressive)"),
  scenario_summary(manta_eval_2025, cull_eval_2025, "sug_Tow_P95", "avoid_Tow_P95", "P95 (Conservative)")
)

kable(consolidated, digits = 2, format.args = list(big.mark = ","), 
      caption = "Consolidated Scenario Comparison: Manta Tow + Cull Dive (All Top 30 Reefs, 2025)")
```

### Confusion Matrices

```{r confusion_matrices, fig.width=12, fig.height=5}
# Function to build and plot a confusion matrix
plot_confusion_matrix <- function(predicted_suggested, observed_presence, title) {
  cm_df <- data.frame(
    Predicted = factor(ifelse(predicted_suggested, "Suggested", "Avoided"), levels = c("Suggested", "Avoided")),
    Observed  = factor(ifelse(observed_presence, "COTS Present", "COTS Absent"), levels = c("COTS Present", "COTS Absent"))
  )
  cm_counts <- cm_df %>%
    count(Predicted, Observed) %>%
    group_by() %>%
    mutate(pct = n / sum(n) * 100)
  
  ggplot(cm_counts, aes(x = Predicted, y = Observed, fill = n)) +
    geom_tile(color = "white", linewidth = 2) +
    geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)") ), size = 5, fontface = "bold") +
    scale_fill_gradient(low = "#f0f4ff", high = "#3366cc") +
    theme_minimal(base_size = 14) +
    labs(title = title, x = "SDM Prediction", y = "Observed", fill = "Count") +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      panel.grid = element_blank()
    )
}

# Manta Tow Confusion Matrices
p1 <- plot_confusion_matrix(
  manta_eval_2025$sug_Tow_P88, manta_eval_2025$presence_any,
  "Manta Tow — P88 (Aggressive)"
)
p2 <- plot_confusion_matrix(
  manta_eval_2025$sug_Tow_P95, manta_eval_2025$presence_any,
  "Manta Tow — P95 (Conservative)"
)

gridExtra::grid.arrange(p1, p2, ncol = 2)

# Cull Dive Confusion Matrices (is_non_problematic as "COTS Absent" proxy)
p3 <- plot_confusion_matrix(
  !cull_eval_2025$avoid_Tow_P88, !cull_eval_2025$is_non_problematic,
  "Cull Dive — P88 (Aggressive)"
)
p4 <- plot_confusion_matrix(
  !cull_eval_2025$avoid_Tow_P95, !cull_eval_2025$is_non_problematic,
  "Cull Dive — P95 (Conservative)"
)

gridExtra::grid.arrange(p3, p4, ncol = 2)
```

## 5. Step D: Visual Mapping

Below are maps for each reef showing the underlying 2025 SDM probability. The operational units displayed are the **100m Buffered Tow Regions**, categorized by the **95th Percentile (P95) Threshold** (Red = Suggested, Grey = Avoided). Individual manta tows and cull dive locations are also plotted to show what falls inside the suggested operational areas.

```{r visual_mapping, results='asis', fig.width=10, fig.height=7}
plot_reef_map <- function(r, tow_sf, dive_sf, region_sf, reef_name, p_min, p_max) {
  # Bounding box around features
  bb_t <- if(nrow(tow_sf) > 0) st_bbox(tow_sf) else NULL
  bb_d <- if(nrow(dive_sf) > 0) st_bbox(dive_sf) else NULL
  
  xmin <- min(c(bb_t["xmin"], bb_d["xmin"]), na.rm=TRUE)
  ymin <- min(c(bb_t["ymin"], bb_d["ymin"]), na.rm=TRUE)
  xmax <- max(c(bb_t["xmax"], bb_d["xmax"]), na.rm=TRUE)
  ymax <- max(c(bb_t["ymax"], bb_d["ymax"]), na.rm=TRUE)
  
  pad <- 2000 # 2km padding
  ext_map <- terra::ext(xmin - pad, xmax + pad, ymin - pad, ymax + pad)
  
  rc <- terra::crop(r, ext_map)
  
  # Scale aggregate factor to keep cell count manageable (~10,000 pixels) for fast plotting
  fact <- ceiling(sqrt(terra::ncell(rc) / 10000))
  if(fact > 1) {
    rc <- terra::aggregate(rc, fact = fact, fun = "mean", na.rm = TRUE)
  }
  
  rdf <- as.data.frame(rc, xy = TRUE, na.rm = TRUE)
  names(rdf)[3] <- "prob"
  
  # Determine if region is suggested based on P95 threshold
  region_sf <- region_sf %>%
    mutate(is_suggested = !is.na(reg_p95) & reg_p95 >= thresh_2025_p95)
  
  gg <- ggplot() +
    geom_raster(data = rdf, aes(x = x, y = y, fill = prob)) +
    # Plot Operational Zones
    geom_sf(data = region_sf, aes(color = is_suggested), fill = NA, linewidth = 1.2, linetype = "dashed") +
    scale_color_manual(values = c("TRUE" = "red", "FALSE" = "grey70")) +
    # Plot Tows and Dives
    geom_sf(data = tow_sf, color = "black", linewidth = 0.5) +
    geom_sf(data = dive_sf, shape = 21, fill = "white", color = "black", size = 2) +
    scale_fill_viridis_c(
      option = "C", 
      na.value = NA, 
      limits = c(p_min, p_max), 
      oob = scales::squish
    ) +
    theme_minimal() +
    labs(
      title = paste("Operational Zonal Map:", reef_name),
      subtitle = paste0("Red Dashed = Suggested 100m Zone (P95), Grey Dashed = Avoided Zone\nScale clipped to ", p_min, "-", p_max),
      fill  = "2025 SDM Prob",
      color = "Suggested Zone"
    )
  return(gg)
}

# Generate plots for each reef
cull_sf_2025 <- cull_sf %>% filter(as.Date(SurveyDate) >= as.Date("2025-01-01"))

for(r_name in reef_select) {
  t_sf <- manta_eval_2025 %>% filter(ReefName == r_name)
  d_sf <- cull_sf_2025 %>% filter(ReefName == r_name)
  r_sf <- tow_regions %>% filter(ReefName == r_name)
  
  if(nrow(t_sf) > 0 || nrow(d_sf) > 0) {
    message(paste("Processing map for:", r_name))
    cat("\n### ", r_name, "\n")
    p <- plot_reef_map(prob_year2025, t_sf, d_sf, r_sf, r_name, p_min_2025, p_max_2025)
    print(p)
    cat("\n")
  }
}
```
