---
title: "ValidateSDM"
format: html
editor: visual
---

## Workflow logic

1.  Need to read in prediction raster and culling and manta tow data along with COTS cull sites

    -   prediction raster `outputs/COTS_prob_problem.tif`

        -   resample to 50m if smaller

    -   Culling data `dat.cull <- read_excel("data/260201_COTS-Cull-Data-Ewels.xlsx", sheet = "Cull")`

    -   Manta tow `dat.manta <- read_csv("data/COTS Program  Manta Tow Data-2026-02-04.csv")`

    -   Cull SIte spatial layer `Eotr_CotsCullSites_2025_11_19.kmz`

        -   should join to `CullSIteName` to `dat.cull` to make it an sf object

2.  Select reefs: Tongue Reef (16-026), Batt Reef (16-029), Arlington Reef (16-064), (Big) Broadhurst Reef (No 1) (18-100a), Britomart Reef (18-024), Sudbury Reef (No 1) (17-001a), Trunk Reef (18-027), Bramble Reef (18-029)

    Develop a method for assigning manta tows to a habitat value.

    -   Connect start and stop as lines and create a 20m buffer around the line

    -   Calculate average, max, 80thperc probability for that tow path

    -   Set threshold (to be tuned) to determine whether that tow would have been conducted or not

    -   Plot by reef

    -   Need some rule to keep mostly continuous path (i.e if there is only 1-2 tows absent it should retain them to only drop whole habitat chunks

    -   For each reef calculate the total tows suggested, tows avoided,

        -   the hit rate of tows that had scars or cots

        -   the proportion/numer of probelmatic COTS culling dives (i.e \>0.02) that were captured/missed - either use overlay of COTS cull sites(polygon) or use centroid (from culling data

        -   the proportion/number of non-probleamtic or 0 cull dives taht would have been voided

```{r}
# ============================================================
# COTS prediction raster validation on selected reefs
# ============================================================

# ---- Packages ----
suppressPackageStartupMessages({
  library(terra)
  library(sf)
  library(dplyr)
  library(readxl)
  library(readr)
  library(lubridate)
  library(ggplot2)
  library(stringr)
})

# ---- Inputs ----
pred_path   <- "outputs/COTS_prob_problem.tif"
cull_xlsx   <- "data/260201_COTS-Cull-Data-Ewels.xlsx"
manta_csv   <- "data/COTS Program  Manta Tow Data-2026-02-04.csv"
cull_kmz    <- "data/Eotr_CotsCullSites_2025_11_19.kmz"

out_dir <- "validation_outputs"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# ---- Selected reefs (edit as needed to match your tables) ----
reef_select <- c(
  "Tongue Reef (16-026)",
  "Batt Reef (16-029)",
  "Arlington Reef (16-064)",
  "(Big) Broadhurst Reef (No 1) (18-100a)",
  "Britomart Reef (18-024)",
  "Sudbury Reef (No 1) (17-001a)",
  "Trunk Reef (18-027)",
  "Bramble Reef (18-029)"
)

# ============================================================
# 1) Helper functions
# ============================================================

# Resample raster to 50m if finer than 50m
resample_to_50m_if_finer <- function(r, target_res = 50) {
  rs <- terra::res(r)
  if (any(is.na(rs))) return(r)

  # "smaller" cell size => finer than target => resample up to target
  if (rs[1] < target_res || rs[2] < target_res) {
    message(sprintf("Raster resolution is %.2fm x %.2fm; resampling to %dm.",
                    rs[1], rs[2], target_res))
    ext_r <- terra::ext(r)
    crs_r <- terra::crs(r)
    template <- terra::rast(
      extent = ext_r,
      crs    = crs_r,
      resolution = target_res
    )
    r2 <- terra::resample(r, template, method = "bilinear")
    return(r2)
  } else {
    message(sprintf("Raster resolution is %.2fm x %.2fm; no resampling needed.",
                    rs[1], rs[2]))
    return(r)
  }
}

# Read KMZ robustly: unzip to temp and read the KML inside.
read_kmz_as_sf <- function(kmz_path) {
  td <- tempfile("kmz_")
  dir.create(td)
  utils::unzip(kmz_path, exdir = td)

  kml_files <- list.files(td, pattern = "\\.kml$", full.names = TRUE, recursive = TRUE)
  if (length(kml_files) == 0) stop("No .kml found inside KMZ.")

  # Some KMZs have multiple KMLs; pick the first
  kml_path <- kml_files[1]
  layers <- tryCatch(st_layers(kml_path), error = function(e) NULL)

  if (!is.null(layers) && length(layers$name) > 0) {
    # Try to read all layers and then you can filter by geometry type
    sfs <- lapply(layers$name, function(lyr) {
      tryCatch(st_read(kml_path, layer = lyr, quiet = TRUE), error = function(e) NULL)
    })
    sfs <- Filter(Negate(is.null), sfs)
    out <- do.call(rbind, sfs)
  } else {
    out <- st_read(kml_path, quiet = TRUE)
  }

  out
}

# Build lines from start/stop lon/lat columns
make_tow_lines <- function(df, lon1, lat1, lon2, lat2, crs_ll = 4326) {
  stopifnot(all(c(lon1, lat1, lon2, lat2) %in% names(df)))

  p1 <- st_as_sf(df, coords = c(lon1, lat1), crs = crs_ll, remove = FALSE)
  p2 <- st_as_sf(df, coords = c(lon2, lat2), crs = crs_ll, remove = FALSE)

  # st_as_sf keeps coords in geometry; we want line from coords explicitly
  c1 <- st_coordinates(p1)
  c2 <- st_coordinates(p2)

  lines <- lapply(seq_len(nrow(df)), function(i) {
    st_linestring(rbind(c1[i,], c2[i,]))
  })

  st_sf(df, geometry = st_sfc(lines, crs = crs_ll))
}

# Extract mean/max/p80 of raster within polygons; returns df keyed by ID
extract_poly_stats <- function(r, polys, id_col = "tow_id", p = 0.8) {
  # terra::extract for polygons returns one row per intersected cell -> summarise
  v <- terra::vect(polys)
  ex <- terra::extract(r, v)  # columns: ID, lyr.1 (or raster name)
  val_col <- setdiff(names(ex), "ID")[1]

  out <- ex |>
    dplyr::rename(poly_id = ID, value = !!val_col) |>
    group_by(poly_id) |>
    summarise(
      prob_mean = mean(value, na.rm = TRUE),
      prob_max  = max(value, na.rm = TRUE),
      prob_p80  = as.numeric(quantile(value, probs = p, na.rm = TRUE, names = FALSE)),
      n_cells   = sum(!is.na(value)),
      .groups   = "drop"
    )

  # Join back to original order
  key <- tibble(poly_id = seq_len(nrow(polys)))
  key |> left_join(out, by = "poly_id")
}

# "Fill short gaps" in a logical vector (keep continuity)
# If there are <= max_gap FALSEs between TRUEs, flip them to TRUE.
fill_short_gaps <- function(x, max_gap = 2) {
  if (length(x) == 0) return(x)
  x <- as.logical(x)
  r <- rle(x)
  vals <- r$values
  lens <- r$lengths

  # Identify FALSE runs that are bounded by TRUE runs
  for (i in seq_along(vals)) {
    if (isFALSE(vals[i]) && lens[i] <= max_gap) {
      left_true  <- i > 1 && isTRUE(vals[i - 1])
      right_true <- i < length(vals) && isTRUE(vals[i + 1])
      if (left_true && right_true) vals[i] <- TRUE
    }
  }
  inverse.rle(list(values = vals, lengths = lens))
}

# Plot helper: raster crop around features and plot as ggplot
plot_reef_map <- function(r, tow_sf, dive_sf, reef_name, file_out) {
  bb <- st_bbox(st_union(tow_sf, dive_sf))
  # add padding in metres (assuming projected CRS)
  pad <- 2000
  bb2 <- bb
  bb2[c("xmin","ymin")] <- bb2[c("xmin","ymin")] - pad
  bb2[c("xmax","ymax")] <- bb2[c("xmax","ymax")] + pad

  rc <- terra::crop(r, terra::ext(bb2["xmin"], bb2["xmax"], bb2["ymin"], bb2["ymax"]))

  # Convert raster to df for ggplot
  rdf <- as.data.frame(rc, xy = TRUE, na.rm = FALSE)
  names(rdf)[3] <- "prob"

  gg <- ggplot() +
    geom_raster(data = rdf, aes(x = x, y = y, fill = prob)) +
    geom_sf(data = tow_sf, aes(color = suggested), fill = NA, linewidth = 0.7) +
    geom_sf(data = dive_sf, shape = 21, fill = "white", color = "black", size = 2) +
    coord_sf(crs = st_crs(tow_sf)) +
    scale_fill_viridis_c(option = "C", na.value = NA) +
    theme_minimal(base_size = 12) +
    labs(
      title = reef_name,
      fill  = "Predicted probability",
      color = "Suggested tow"
    )

  ggsave(file_out, gg, width = 10, height = 7, dpi = 300)
  gg
}

```

```{r}
# ============================================================
# 2) Read prediction raster and harmonise CRS / res
# ============================================================

pred <- terra::rast(pred_path)

# If raster is lon/lat, you MUST project before buffering in metres.
# These reefs are in/near Cairns -> MGA Zone 55 is typically appropriate.
# If your raster is already projected in metres, keep it.
if (terra::is.lonlat(pred)) {
  message("Raster is lon/lat; projecting to EPSG:7855 (GDA2020 / MGA zone 55) for metric operations.")
  pred <- terra::project(pred, "EPSG:7855", method = "bilinear")
}
res(pred)
pred <- resample_to_50m_if_finer(pred, target_res = 50)

# ============================================================
# 3) Read tabular data
# ============================================================

dat.cull  <- read_excel(cull_xlsx, sheet = "Cull") |> as.data.frame() |>
  mutate(Total = Cohort1 + Cohort2 + Cohort3 + Cohort4)
dat.manta <- read_csv(manta_csv, show_col_types = FALSE) |> as.data.frame()

# ============================================================
# 4) Column mapping (EDIT THIS SECTION ONCE)
# ============================================================

# ---- Manta tow columns ----
# Required: reef name/id; tow id; start/end lon/lat; an ordering variable (optional but helps)
manta_cols <- list(
  reef      = "ReefName",       # or "Reef_Name" etc
  tow_id    = "CrownOfThornsStarfishSurveillanceId",          # unique id per tow
  date      = "SurveyTime",           # can be Date or POSIXct
  lon_start = "StartLongitude",
  lat_start = "StartLatitude",
  lon_end   = "EndLongitude",
  lat_end   = "EndLatitude",
  scars     = "FeedingScarCountRangeCode",           # indicator or count
  cots      = "CrownOfThornsStarfishCount"            # indicator or count
)

# ---- Cull columns ----
# If you have actual dive lon/lat in the cull table, set these; otherwise leave NULL and use cull sites geometry.
cull_cols <- list(
  reef       = "ReefName",
  cull_site  = "CullSiteName",  # joins to KMZ attribute
  date       = "SurveyDate",
  culls      = "Total",
  time       = "Bottomtime",
  lon        = "Longitude",            # e.g. "Longitude"
  lat        = "Latitude"             # e.g. "Latitude"
)

# ---- Cull sites (KMZ) attribute column containing site name ----
cullsite_name_col <- "Name"  # EDIT if the KMZ uses a different field name

# ============================================================
# 5) Filter to selected reefs
# ============================================================

dat.manta <- dat.manta |>
  mutate(.reef = as.character(.data[[manta_cols$reef]])) |>
  filter(.reef %in% reef_select) |>
  filter(!is.na(StartLatitude), !is.na(EndLatitude))

dat.cull <- dat.cull |>
  mutate(.reef = as.character(.data[[cull_cols$reef]])) |>
  filter(.reef %in% reef_select)

# ============================================================
# 6) Read cull sites (KMZ) and build dive geometries
# ============================================================

cull_sites <- read_kmz_as_sf(cull_kmz)

# Keep polygon-like geometries (sites). If KMZ has points only, this still works.
cull_sites <- cull_sites |>
  mutate(.geom_type = as.character(st_geometry_type(geometry))) |>
  filter(.geom_type %in% c("POLYGON","MULTIPOLYGON","POINT","MULTIPOINT","LINESTRING","MULTILINESTRING")) |>
  select(-.geom_type)

if (!(cullsite_name_col %in% names(cull_sites))) {
  stop(sprintf("Cull site name column '%s' not found in KMZ. Available: %s",
               cullsite_name_col, paste(names(cull_sites), collapse = ", ")))
}

# Project cull sites into raster CRS (metric)
cull_sites <- st_transform(cull_sites, st_crs(pred))

# Build cull "dive" sf: Join to cull_sites by CullSiteName -> use polygon; also create centroid

cull_sf <- dat.cull |>
    mutate(.site = as.character(.data[[cull_cols$cull_site]])) |>
    left_join(
      cull_sites |> st_drop_geometry() |> mutate(.site = as.character(.data[[cullsite_name_col]])),
      by = ".site"
    )

  # Attach geometry by joining with cull_sites again (sf join by attribute)
cull_sf <- cull_sf |>
    left_join(
      cull_sites |> mutate(.site = as.character(.data[[cullsite_name_col]])) |> select(.site, geometry),
      by = ".site"
    ) |>
    st_as_sf()

# If polygons, you can work with polygons OR centroids; keep both
cull_sf_cent <- st_centroid(cull_sf) |> mutate(dive_geom = "centroid")
cull_sf_poly <- cull_sf |> mutate(dive_geom = "site_geom")

cull_sf <- bind_rows(cull_sf_poly, cull_sf_cent)

# ============================================================
# 7) Build tow lines, buffer 20 m, extract raster stats
# ============================================================

# Parse date if needed (safe attempt)
if (manta_cols$date %in% names(dat.manta)) {
  dat.manta[[manta_cols$date]] <- suppressWarnings(ymd(dat.manta[[manta_cols$date]]))
}

tow_lines_ll <- make_tow_lines(
  dat.manta,
  lon1 = manta_cols$lon_start, lat1 = manta_cols$lat_start,
  lon2 = manta_cols$lon_end,   lat2 = manta_cols$lat_end,
  crs_ll = 4326
)

# Project to raster CRS for buffering in metres
tow_lines <- st_transform(tow_lines_ll, st_crs(pred))

# Tow buffer
tow_buf <- st_buffer(tow_lines, dist = 20)

# Give each tow a local numeric id for extraction
tow_buf <- tow_buf |>
  mutate(
    tow_row_id = row_number(),
    reef = as.character(.data[[manta_cols$reef]]),
    tow_id = as.character(.data[[manta_cols$tow_id]])
  )

stats <- extract_poly_stats(pred, tow_buf, id_col = "tow_row_id", p = 0.8)
tow_buf <- bind_cols(tow_buf, stats |> select(prob_mean, prob_max, prob_p80, n_cells))

```

```{r}
# 8) Thresholding + continuity rule
# ============================================================

# Choose which metric drives decisions (mean/max/p80)
decision_metric <- "prob_p80"

# Threshold to tune
tow_threshold <- 0.415

tow_buf <- tow_buf |>
  mutate(
    metric = .data[[decision_metric]],
    suggested_raw = metric >= tow_threshold
  )

# Order within reef (and optionally date) to apply continuity
tow_buf <- tow_buf |>
  group_by(reef) |>
  mutate(.order = row_number()) |>
  ungroup()


# Apply "fill short gaps" within each reef and (if available) within day
group_vars <- c("reef")
if (manta_cols$date %in% names(tow_buf)) group_vars <- c(group_vars, manta_cols$date)

tow_buf <- tow_buf |>
  arrange(across(all_of(group_vars)), .order) |>
  group_by(across(all_of(group_vars))) |>
  mutate(
    suggested = fill_short_gaps(suggested_raw, max_gap = 2)
  ) |>
  ungroup()
table(tow_buf$suggested); table(tow_buf$suggested_raw)
# ============================================================
# 9) Hit-rate metrics on manta tows (scars or COTS)
# ============================================================

# Define "positive" tow (edit if your columns differ)
tow_buf <- tow_buf |>
  mutate(
    scars = .data[[manta_cols$scars]],
    cots  = as.numeric(.data[[manta_cols$cots]]),
    positive = scars %in% c("p", "c") | (coalesce(cots, 0) > 0)
  )


# ============================================================
# 10) Cull dive probability + captured/missed by suggested tows
# ============================================================

# Extract probability at each dive geometry
# - For centroid/point: extract at point
# - For polygon: extract mean or max within polygon (use mean here; change if you want max)

get_point_prob <- function(r, pts_sf) {
  v <- terra::vect(pts_sf)
  ex <- terra::extract(r, v)
  val_col <- setdiff(names(ex), "ID")[1]
  ex[[val_col]]
}

# Separate by geometry “type”
cull_cent <- cull_sf |> filter(dive_geom %in% c("point","centroid"))
cull_poly <- cull_sf |> filter(dive_geom == "site_geom")

if (nrow(cull_cent) > 0) {
  cull_cent$prob <- get_point_prob(pred, cull_cent)
}
# if (nrow(cull_poly) > 0) {
#   v <- terra::vect(cull_poly)
#   ex <- terra::extract(pred, v)
#   val_col <- setdiff(names(ex), "ID")[1]
#   cull_poly$prob <- ex |>
#     rename(poly_id = ID, value = !!val_col) |>
#     group_by(poly_id) |>
#     summarise(prob = mean(value, na.rm = TRUE), .groups = "drop") |>
#     pull(prob)
# }

cull_sf <- bind_rows(cull_cent, cull_poly) |>
  mutate(
    reef = as.character(.data[[cull_cols$reef]]),
    culls = as.numeric(.data[[cull_cols$culls]])
  )

# Create per-reef union of suggested tow buffers for capture tests
tow_suggested_union <- tow_buf |>
  filter(suggested) |>
  group_by(reef) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

tow_nonsuggested_union <- tow_buf |>
  filter(!suggested) |>
  group_by(reef) |>
  summarise(geometry = st_union(geometry), .groups = "drop")

crs(tow_suggested_union); crs(cull_sf)

plot(tow_suggested_union[2,])
# Tag each dive as captured/missed by suggested tow coverage
cull_sf <- cull_sf |>
  filter(dive_geom=="site_geom", SurveyDate > "2018-11-01") |>
  mutate(
    captured = lengths(st_intersects(geometry, tow_suggested_union$geometry)) > 0,
    problematic = (culls / Bottomtime) > 0.01
  ) 



```

```{r}



library(purrr)
library(patchwork)
library(scales)
library(ggplot2)

reefs <- intersect(unique(cull_sf$reef), unique(tow_suggested_union$reef))
cull_sf <- cull_sf |>
  mutate(
    captured.fac   = factor(as.character(captured),    levels = c("FALSE","TRUE")),
    problematic.fac = factor(as.character(problematic), levels = c("FALSE","TRUE"))
  )

# Common scales + guides (IDENTICAL for every plot)
col_scale  <- scale_color_discrete(limits = c("FALSE","TRUE"), drop = FALSE, name = "captured")
fill_scale <- scale_fill_discrete( limits = c("FALSE","TRUE"), drop = FALSE, name = "problematic")

common_guides <- guides(
  color = guide_legend(order = 1, override.aes = list(alpha = 1, fill = NA)),
  fill  = guide_legend(order = 2, override.aes = list(alpha = 1))
)


base_theme <- theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 7),
    axis.title = element_blank(),
    panel.grid.minor = element_blank()
  )


plots <- map(reefs, \(rf) {
  u <- tow_suggested_union |> filter(reef == rf)
  c <- cull_sf |> filter(reef == rf)

  ggplot() +
    geom_sf(data = u, fill = NA, linewidth = 0.6) +
    geom_sf(data = c, aes(color = captured, fill = problematic), alpha = 0.8) +
    coord_sf(label_graticule = "SE") +
    scale_x_continuous(breaks = pretty_breaks(n = 3)) +
    scale_y_continuous(breaks = pretty_breaks(n = 3)) +
    col_scale + fill_scale + common_guides +          # <- important: same objects everywhere
    base_theme +
    labs(title = rf, x = NULL, y = NULL)
})

p <- wrap_plots(plots, ncol = 2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

p

ggsave("validation_outputs/Reefplots.png", plot = p, width = 8, height = 12, dpi = 300)

```

```{r}
# ============================================================
# 11) Summaries per reef
# ============================================================

tow_summary <- tow_buf |>
  st_drop_geometry() |>
  group_by(reef) |>
  summarise(
    n_tows_total     = n(),
    n_tows_suggested = sum(suggested, na.rm = TRUE),
    n_tows_avoided   = sum(!suggested, na.rm = TRUE),

    # Confusion matrix counts (suggested = model positive, positive = observed positive)
    TP = sum(suggested &  positive, na.rm = TRUE),  # suggested and truly positive
    FP = sum(suggested & !positive, na.rm = TRUE),  # suggested but truly negative
    FN = sum(!suggested &  positive, na.rm = TRUE), # avoided but truly positive
    TN = sum(!suggested & !positive, na.rm = TRUE), # avoided and truly negative
    
    TP.rate = TP/n_tows_suggested,  # suggested and truly positive
    FP.rate = FP/n_tows_suggested,  # suggested but truly negative
    FN.rate = FN/n_tows_avoided, # avoided but truly positive
    TN.rate = TN/n_tows_avoided,

    # Optional: totals
    n_positive = sum(positive, na.rm = TRUE),
    n_negative = sum(!positive, na.rm = TRUE),

    # Optional: derived rates (drop if you only want counts)
    sensitivity = ifelse((TP + FN) > 0, TP / (TP + FN), NA_real_),  # recall / TPR
    specificity = ifelse((TN + FP) > 0, TN / (TN + FP), NA_real_),  # TNR
    precision   = ifelse((TP + FP) > 0, TP / (TP + FP), NA_real_),  # PPV
    npv         = ifelse((TN + FN) > 0, TN / (TN + FN), NA_real_),  # NPV
    accuracy    = ifelse(n() > 0, (TP + TN) / n(), NA_real_),

    .groups = "drop"
  )
# Cull summaries:
# - problematic dives (prob>0.02): captured vs missed
# - non-problematic OR zero-cull dives avoided: i.e., NOT captured by suggested tow coverage
cull_summary <- cull_sf |>
  st_drop_geometry() |>
  group_by(reef) |>
  summarise(
    n_dives_total = n(),
    n_problematic = sum(problematic, na.rm = TRUE),
    n_problematic_captured = sum(problematic & captured, na.rm = TRUE),
    n_problematic_missed   = sum(problematic & !captured, na.rm = TRUE),
    prop_problematic_captured = ifelse(sum(problematic, na.rm = TRUE) > 0,
                                       sum(problematic & captured, na.rm = TRUE) / sum(problematic, na.rm = TRUE),
                                       NA_real_),
     # Total COTS missed due to missed problematic sites
    cots_missed_problematic = sum(coalesce(culls, 0)[problematic & !captured], 
                                  na.rm = TRUE),
    cpue_missed_problematic = {
  denom <- sum(coalesce(Bottomtime, 0)[problematic & !captured], na.rm = TRUE)
  ifelse(denom > 0,
         sum(coalesce(culls, 0)[problematic & !captured], na.rm = TRUE) / denom,
         NA_real_)
},
    n_nonprob_or_zero = sum((!problematic) | (coalesce(culls, 0) == 0), na.rm = TRUE),
    n_nonprob_or_zero_avoided = sum((((!problematic) | (coalesce(culls, 0) == 0)) & (!captured)), na.rm = TRUE),
    prop_nonprob_or_zero_avoided = ifelse(sum((!problematic) | (coalesce(culls, 0) == 0), na.rm = TRUE) > 0,
                                          sum((((!problematic) | (coalesce(culls, 0) == 0)) & (!captured)), na.rm = TRUE) /
                                            sum((!problematic) | (coalesce(culls, 0) == 0), na.rm = TRUE),
                                          NA_real_),
    .groups = "drop"
  )

summary_all <- tow_summary |>
  left_join(cull_summary, by = "reef") |>
  arrange(reef)

write_csv(summary_all, file.path(out_dir, "reef_validation_summary.csv"))
print(summary_all)

```

```{r}
# install.packages(c("maptiles", "tidyterra"))
library(maptiles)
library(tidyterra)
library(sf)
library(dplyr)
library(ggplot2)
library(scales)

tow_anno <- summary_all |>
  mutate(
    anno = sprintf(
      "Tows: %d (S %d / A %d)\n TPR %s  PPV %s\nPcap %s  CPUEmiss %s  Avoid0 %s",
      n_tows_total, n_tows_suggested, n_tows_avoided,
      ifelse(is.na(sensitivity), "NA", percent(sensitivity, accuracy = 1)),
      ifelse(is.na(precision),   "NA", percent(precision, accuracy = 1)),
      ifelse(is.na(prop_problematic_captured), "NA", percent(prop_problematic_captured, accuracy = 1)),
      ifelse(is.na(cpue_missed_problematic),   "NA", sprintf("%.3f", cpue_missed_problematic)),
      ifelse(is.na(prop_nonprob_or_zero_avoided), "NA", percent(prop_nonprob_or_zero_avoided, accuracy = 1))
    )
  ) |>
  select(reef, anno)

library(viridisLite)

# pick two colours from the same palette
cols <- viridis(256)
col_false <- cols[90]   # green-ish (tweak index if you want)
col_true  <- cols[5]  # red/purple (tweak index if you want)

pal_tf <- c("FALSE" = col_false, "TRUE" = col_true)


for (rf in unique(tow_buf$reef)) {

  u <- tow_suggested_union |> filter(reef == rf)
  nu <- tow_nonsuggested_union |> filter(reef == rf)
  c <- cull_sf |> filter(reef == rf, dive_geom == "site_geom")

  if (nrow(u) == 0) next

  lab <- tow_anno |> filter(reef == rf) |> pull(anno)
  if (length(lab) == 0) lab <- "No tow summary"
  # --- transform to Web Mercator for map tiles ---
  u3857  <- st_transform(u, 3857)
  nu3857 <- st_transform(nu, 3857)
  c3857  <- st_transform(c, 3857)
  # bbox with padding (metres in EPSG:3857)
  bb <- st_bbox(st_union(u3857, c3857))
  pad <- 2000
  bb[c("xmin","ymin")] <- bb[c("xmin","ymin")] - pad
  bb[c("xmax","ymax")] <- bb[c("xmax","ymax")] + pad
  bb_sfc <- st_as_sfc(bb)

  img <- maptiles::get_tiles(bb_sfc, provider = "Esri.WorldImagery",
                           zoom = 13, crop = TRUE,
                           cachedir = file.path(out_dir, "tiles_cache"))

  
  p_rf <-
    ggplot() +
    tidyterra::geom_spatraster_rgb(data=img) +
    geom_sf(data = u3857, fill = NA, color = "blue", linewidth = 0.6) +
    # geom_sf(data = nu3857, fill = NA, color = "purple", linewidth = 0.6) +
    geom_sf(data = c3857, aes(color = captured, fill = problematic), alpha = 0.8) +
    coord_sf(crs = 3857, datum = NA) +
    # geom_sf(data = u, fill = NA,color="blue",linewidth = 0.6) +
    # # geom_sf(data = nu, fill = NA,color="purple", linewidth = 0.6) +
    # geom_sf(data = c, aes(color = captured, fill = problematic), alpha = 0.8) +
    # coord_sf(label_graticule = "SE") +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 3)) +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) +
    # keep legend consistent if you want (optional)
    scale_color_manual(values = pal_tf, limits = c("FALSE","TRUE"), drop = FALSE, name = "captured") +
    scale_fill_manual(values = pal_tf, limits = c("FALSE","TRUE"), drop = FALSE, name = "problematic") +
    # guides(color = "none", fill = "none") +  # <- per-reef files: remove legend
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      axis.text.y = element_text(size = 7),
      panel.grid.minor = element_blank(), legend.position = "bottom"
    ) +
    labs(title = rf, x = NULL, y = NULL) +
    # annotation box in plot corner
    annotate(
      "label",
      x = -Inf, y = Inf, hjust = -0.05, vjust = 1.1,
      label = lab,
      size = 3, 
    )

  plot_path <- file.path(out_dir, paste0(stringr::str_replace_all(rf, "[^A-Za-z0-9]+", "_"), "_map.png"))
  ggsave(plot_path, plot = p_rf, width = 8, height = 6, dpi = 300)
}

```
